---
title: "Additional tests IF, GRAPH etc"
output: pdf_document
date: "2026-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
source("prepare_data.R")
```

# NEW IF - addedmore features

```{r}
library(isotree)

# 1. Basket-level features for IF
if_train <- basket_data %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    max_time,
    min_time,
    range_time,
    cv_time,
    price_per_second,
    items_per_second,
    max_price,
    min_price,
    sd_price,
    max_price_share,
    top_3_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    items_per_department,
    price_deviation,
    time_deviation,
    long_pause_count
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(sd_time_per_item, na.rm = TRUE)),
    sd_price = replace_na(sd_price, median(sd_price, na.rm = TRUE)),
    cv_time = replace_na(cv_time, median(cv_time, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(max_price_share, na.rm = TRUE)),
    top_3_price_share = replace_na(top_3_price_share, median(top_3_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(dominant_department_share, na.rm = TRUE)),
    items_per_department = replace_na(items_per_department, median(items_per_department, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(time_deviation, na.rm = TRUE))
  )

# 2. Matrix for IF
X_train_raw <- if_train %>%
  select(-transaction_id)

X_train <- scale(X_train_raw)

# 3. Fit IF on supermarket basket data
if_model <- isolation.forest(
  as.matrix(X_train),
  ntrees = 300,
  sample_size = 512,
  ndim = 1,
  nthreads = 1
)

# 4. Get anomaly scores on supermarket
if_train$if_score <- predict(
  if_model,
  as.matrix(X_train),
  type = "score"
)

# 5. Since known fraud rate is about 15%, flag top 15%
if_threshold <- quantile(if_train$if_score, 0.85, na.rm = TRUE)

if_train <- if_train %>%
  mutate(
    if_flag = if_score >= if_threshold
  ) %>%
  arrange(desc(if_score))

table(if_train$if_flag)

# Top suspicious supermarket transactions according to IF
top_if_train <- if_train %>%
  arrange(desc(if_score)) %>%
  slice_head(n = 150)

View(top_if_train)

write.csv(
  top_if_train,
  "~/Desktop/top_if_train.csv",
  row.names = FALSE
)

top_if_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_time_per_item,
    sd_time_per_item,
    range_time,
    cv_time,
    price_per_second,
    max_price,
    sd_price,
    max_price_share,
    top_3_price_share,
    unique_departments,
    dominant_department_share,
    items_per_department,
    long_pause_count,
    if_score
  ) %>%
  head(20)
```

Слишком тяготеет к маленьким корзинам на основании старой расширенной версии IF, с точки зрения экономического эффекта для бизнеса - не эффективен.

Новый IF выглядит сильным как detector аномального поведения, но он:

снижает средний чек
снижает средний размер корзины
значит потенциально режет именно тот тип транзакций, который дает много денег и много товаров




# pure graph + addition of markov chain/models
# MARKOV MODEL ON DEPARTMENT TRANSITIONS

```{r}

# 1. Read raw dataset with item order preserved
lines <- readLines("~/Desktop/supermarket case/supermarket.csv", warn = FALSE)

supermarket_raw <- tibble(
  transaction_id = 1:length(lines),
  basket = lines
)

dept_map <- c(
  "1" = "Bakery & Pastry",
  "2" = "Beer & Wine",
  "3" = "Books & Magazines",
  "4" = "Candy & Chips",
  "5" = "Care & Hygiene",
  "6" = "Cereals & Spreads",
  "7" = "Cheese & Tapas",
  "8" = "Dairy & Eggs",
  "9" = "Freezer",
  "10" = "Fruit & Vegetables",
  "11" = "Household & Pet",
  "12" = "Meat & Fish",
  "13" = "Pasta & Rice",
  "14" = "Salads & Meals",
  "15" = "Sauces & Spices",
  "16" = "Soda & Juices",
  "17" = "Special Diet",
  "18" = "Vegetarian & Vegan"
)

long_seq <- supermarket_raw %>%
  separate_rows(basket, sep = ",") %>%
  group_by(transaction_id) %>%
  mutate(item_order = row_number()) %>%
  ungroup() %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    !is.na(department_name),
    time >= 0,
    price >= 0
  )

# 2. Build transitions inside each basket
transitions <- long_seq %>%
  arrange(transaction_id, item_order) %>%
  group_by(transaction_id) %>%
  mutate(
    next_department = lead(department_name)
  ) %>%
  ungroup() %>%
  filter(!is.na(next_department)) %>%
  select(transaction_id, from_dept = department_name, to_dept = next_department)

# 3. Count transitions on supermarket
transition_counts <- transitions %>%
  count(from_dept, to_dept, name = "transition_count")

from_counts <- transitions %>%
  count(from_dept, name = "from_count")

# 4. Transition probabilities
transition_probs <- transition_counts %>%
  left_join(from_counts, by = "from_dept") %>%
  mutate(
    transition_prob = transition_count / from_count,
    transition_neglog = -log(transition_prob)
  )

# 5. Basket-level Markov score
markov_train <- transitions %>%
  left_join(
    transition_probs %>%
      select(from_dept, to_dept, transition_prob, transition_neglog),
    by = c("from_dept", "to_dept")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    markov_score = mean(transition_neglog, na.rm = TRUE),
    avg_transition_prob = mean(transition_prob, na.rm = TRUE),
    n_transitions = n(),
    .groups = "drop"
  )

# 6. Join to basket_data
basket_markov <- basket_data %>%
  left_join(markov_train, by = "transaction_id") %>%
  mutate(
    markov_score = replace_na(markov_score, 0),
    avg_transition_prob = replace_na(avg_transition_prob, 0),
    n_transitions = replace_na(n_transitions, 0)
  ) %>%
  arrange(desc(markov_score))

# 7. Top suspicious baskets by Markov score
top_150_markov_train <- basket_markov %>%
  slice_head(n = 150)

top_150_markov_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    unique_departments,
    markov_score,
    avg_transition_prob,
    n_transitions
  ) %>%
  head(20)

write_csv(
  top_150_markov_train,
  "~/Desktop/top_150_markov_train.csv"
)
```

# CASE 9: IF + markov train

```{r}
lines <- readLines("~/Desktop/supermarket case/supermarket.csv", warn = FALSE)

supermarket_raw <- tibble(
  transaction_id = 1:length(lines),
  basket = lines
)

dept_map <- c(
  "1" = "Bakery & Pastry",
  "2" = "Beer & Wine",
  "3" = "Books & Magazines",
  "4" = "Candy & Chips",
  "5" = "Care & Hygiene",
  "6" = "Cereals & Spreads",
  "7" = "Cheese & Tapas",
  "8" = "Dairy & Eggs",
  "9" = "Freezer",
  "10" = "Fruit & Vegetables",
  "11" = "Household & Pet",
  "12" = "Meat & Fish",
  "13" = "Pasta & Rice",
  "14" = "Salads & Meals",
  "15" = "Sauces & Spices",
  "16" = "Soda & Juices",
  "17" = "Special Diet",
  "18" = "Vegetarian & Vegan"
)

long_data <- supermarket_raw %>%
  separate_rows(basket, sep = ",") %>%
  group_by(transaction_id) %>%
  mutate(item_order = row_number()) %>%
  ungroup() %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  )

long_data_clean <- long_data %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    !is.na(department_name),
    time >= 0,
    price >= 0
  ) %>%
  mutate(
    price_per_second = ifelse(time > 0, price / time, NA_real_)
  )

# basket data for IF
basket_data <- long_data_clean %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(sum(time, na.rm = TRUE) > 0,
                              sum(price, na.rm = TRUE) / sum(time, na.rm = TRUE),
                              NA_real_),
    max_price = max(price, na.rm = TRUE),
    max_price_share = ifelse(sum(price, na.rm = TRUE) > 0,
                             max(price, na.rm = TRUE) / sum(price, na.rm = TRUE),
                             NA_real_),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

# pure IF train
if_train <- basket_data %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    max_price,
    max_price_share,
    unique_departments
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(price_per_second, na.rm = TRUE))
  )

X_train_raw <- if_train %>%
  select(-transaction_id)

X_train <- scale(X_train_raw)

if_model <- isolation.forest(
  as.matrix(X_train),
  ntrees = 300,
  sample_size = 512,
  ndim = 1,
  nthreads = 1
)

if_train$if_score <- predict(
  if_model,
  as.matrix(X_train),
  type = "score"
)

# Markov train
transitions <- long_data_clean %>%
  arrange(transaction_id, item_order) %>%
  group_by(transaction_id) %>%
  mutate(next_department = lead(department_name)) %>%
  ungroup() %>%
  filter(!is.na(next_department)) %>%
  select(transaction_id,
         from_dept = department_name,
         to_dept = next_department)

transition_counts <- transitions %>%
  count(from_dept, to_dept, name = "transition_count")

from_counts <- transitions %>%
  count(from_dept, name = "from_count")

transition_probs <- transition_counts %>%
  left_join(from_counts, by = "from_dept") %>%
  mutate(
    transition_prob = transition_count / from_count,
    transition_neglog = -log(transition_prob)
  )

markov_train <- transitions %>%
  left_join(
    transition_probs %>%
      select(from_dept, to_dept, transition_prob, transition_neglog),
    by = c("from_dept", "to_dept")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    markov_score = mean(transition_neglog, na.rm = TRUE),
    avg_transition_prob = mean(transition_prob, na.rm = TRUE),
    n_transitions = n(),
    .groups = "drop"
  )

basket_if_markov <- basket_data %>%
  left_join(
    if_train %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  left_join(
    markov_train,
    by = "transaction_id"
  ) %>%
  mutate(
    markov_score = replace_na(markov_score, 0),
    avg_transition_prob = replace_na(avg_transition_prob, 0),
    n_transitions = replace_na(n_transitions, 0),
    if_score_01 = percent_rank(if_score),
    markov_score_01 = percent_rank(markov_score),
    combined_im_score = 0.6 * if_score_01 + 0.4 * markov_score_01
  ) %>%
  arrange(desc(combined_im_score))

top_150_if_markov_train <- basket_if_markov %>%
  slice_head(n = 150)

top_150_if_markov_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_time_per_item,
    if_score,
    markov_score,
    if_score_01,
    markov_score_01,
    combined_im_score
  ) %>%
  head(20)

write.csv(
  top_150_if_markov_train,
  "~/Desktop/top_150_if_markov_train.csv",
  row.names = FALSE
)
```


# APPLY ON CASE 9
```{r}
case9_lines <- readLines("~/Desktop/case9.csv", warn = FALSE)

case9_raw <- tibble(
  raw_line = case9_lines
) %>%
  mutate(
    transaction_id = stringr::str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = stringr::str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case9_long <- case9_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  group_by(transaction_id) %>%
  mutate(item_order = row_number()) %>%
  ungroup() %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    !is.na(department_name),
    time >= 0,
    price >= 0
  )

case9_basket <- case9_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(sum(time, na.rm = TRUE) > 0,
                              sum(price, na.rm = TRUE) / sum(time, na.rm = TRUE),
                              NA_real_),
    max_price = max(price, na.rm = TRUE),
    max_price_share = ifelse(sum(price, na.rm = TRUE) > 0,
                             max(price, na.rm = TRUE) / sum(price, na.rm = TRUE),
                             NA_real_),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

# case9 IF score
if_case9 <- case9_basket %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    max_price,
    max_price_share,
    unique_departments
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(if_train$sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(if_train$price_per_second, na.rm = TRUE))
  )

X_case9_raw <- if_case9 %>%
  select(-transaction_id)

X_case9 <- scale(
  X_case9_raw,
  center = attr(X_train, "scaled:center"),
  scale = attr(X_train, "scaled:scale")
)

if_case9$if_score <- predict(
  if_model,
  as.matrix(X_case9),
  type = "score"
)

# case9 Markov score
case9_transitions <- case9_long %>%
  arrange(transaction_id, item_order) %>%
  group_by(transaction_id) %>%
  mutate(next_department = lead(department_name)) %>%
  ungroup() %>%
  filter(!is.na(next_department)) %>%
  select(transaction_id,
         from_dept = department_name,
         to_dept = next_department)

case9_markov <- case9_transitions %>%
  left_join(
    transition_probs %>%
      select(from_dept, to_dept, transition_prob, transition_neglog),
    by = c("from_dept", "to_dept")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    markov_score = mean(transition_neglog, na.rm = TRUE),
    avg_transition_prob = mean(transition_prob, na.rm = TRUE),
    n_transitions = n(),
    .groups = "drop"
  )

# final IF + Markov
case9_im <- case9_basket %>%
  left_join(
    if_case9 %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  left_join(
    case9_markov,
    by = "transaction_id"
  ) %>%
  mutate(
    markov_score = replace_na(markov_score, 0),
    avg_transition_prob = replace_na(avg_transition_prob, 0),
    n_transitions = replace_na(n_transitions, 0),
    if_score_01 = percent_rank(if_score),
    markov_score_01 = percent_rank(markov_score),
    combined_im_score = 0.6 * if_score_01 + 0.4 * markov_score_01
  ) %>%
  arrange(desc(combined_im_score))

top_150_if_markov_case9 <- case9_im %>%
  slice_head(n = 150)

summary(top_150_if_markov_case9$if_score)
summary(top_150_if_markov_case9$markov_score)

writeLines(as.character(top_150_if_markov_case9$transaction_id))

write.csv(
  top_150_if_markov_case9,
  "~/Desktop/top_150_if_markov_case9.csv",
  row.names = FALSE
)
```

# CASE 10: IF + GRAPH + SMALL MARKOV TRAIN ON SUPERMARKET

```{r}
# weights:
# IF      0.55
# GRAPH   0.35
# MARKOV  0.10

basket_igm_train <- basket_data %>%
  left_join(
    if_train %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  left_join(
    graph_train %>% select(transaction_id, graph_score),
    by = "transaction_id"
  ) %>%
  left_join(
    markov_train %>% select(transaction_id, markov_score),
    by = "transaction_id"
  ) %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    markov_score = replace_na(markov_score, 0),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    markov_score_01 = percent_rank(markov_score),
    combined_igm_score =
      0.55 * if_score_01 +
      0.35 * graph_score_01 +
      0.10 * markov_score_01
  ) %>%
  arrange(desc(combined_igm_score))

top_150_if_graph_markov_train <- basket_igm_train %>%
  slice_head(n = 150)

top_150_if_graph_markov_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_time_per_item,
    if_score,
    graph_score,
    markov_score,
    if_score_01,
    graph_score_01,
    markov_score_01,
    combined_igm_score
  ) %>%
  head(20)

summary(top_150_if_graph_markov_train$if_score)
summary(top_150_if_graph_markov_train$graph_score)
summary(top_150_if_graph_markov_train$markov_score)
summary(top_150_if_graph_markov_train$total_items)
summary(top_150_if_graph_markov_train$total_price)

write.csv(
  top_150_if_graph_markov_train,
  "~/Desktop/top_150_if_graph_markov_train.csv",
  row.names = FALSE
)
```

# APPLY ON CASE 10

```{r}
# IF + GRAPH + SMALL MARKOV
# weights:
# IF      0.55
# GRAPH   0.35
# MARKOV  0.10

case10_lines <- readLines("~/Desktop/case10.csv", warn = FALSE)

case10_raw <- tibble(
  raw_line = case10_lines
) %>%
  mutate(
    transaction_id = stringr::str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = stringr::str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case10_long <- case10_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  group_by(transaction_id) %>%
  mutate(item_order = row_number()) %>%
  ungroup() %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    !is.na(department_name),
    time >= 0,
    price >= 0
  )

# basket features

case10_basket <- case10_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(sum(time, na.rm = TRUE) > 0,
                              sum(price, na.rm = TRUE) / sum(time, na.rm = TRUE),
                              NA_real_),
    max_price = max(price, na.rm = TRUE),
    max_price_share = ifelse(sum(price, na.rm = TRUE) > 0,
                             max(price, na.rm = TRUE) / sum(price, na.rm = TRUE),
                             NA_real_),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

# IF score

if_case10 <- case10_basket %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    max_price,
    max_price_share,
    unique_departments
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(if_train$sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(if_train$price_per_second, na.rm = TRUE))
  )

X_case10_raw <- if_case10 %>%
  select(-transaction_id)

X_case10 <- scale(
  X_case10_raw,
  center = attr(X_train, "scaled:center"),
  scale = attr(X_train, "scaled:scale")
)

if_case10$if_score <- predict(
  if_model,
  as.matrix(X_case10),
  type = "score"
)

# GRAPH score
# requires pair_conditional from graph train

case10_dept_data <- case10_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case10_pairs <- case10_dept_data %>%
  inner_join(case10_dept_data, by = "transaction_id", relationship = "many-to-many") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

case10_graph <- case10_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

# MARKOV score
# requires transition_probs from markov train

case10_transitions <- case10_long %>%
  arrange(transaction_id, item_order) %>%
  group_by(transaction_id) %>%
  mutate(next_department = lead(department_name)) %>%
  ungroup() %>%
  filter(!is.na(next_department)) %>%
  select(
    transaction_id,
    from_dept = department_name,
    to_dept = next_department
  )

case10_markov <- case10_transitions %>%
  left_join(
    transition_probs %>%
      select(from_dept, to_dept, transition_prob, transition_neglog),
    by = c("from_dept", "to_dept")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    markov_score = mean(transition_neglog, na.rm = TRUE),
    avg_transition_prob = mean(transition_prob, na.rm = TRUE),
    n_transitions = n(),
    .groups = "drop"
  )

# FINAL COMBINED SCORE

case10_combo <- case10_basket %>%
  left_join(
    if_case10 %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  left_join(
    case10_graph %>% select(transaction_id, graph_score),
    by = "transaction_id"
  ) %>%
  left_join(
    case10_markov %>% select(transaction_id, markov_score),
    by = "transaction_id"
  ) %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    markov_score = replace_na(markov_score, 0),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    markov_score_01 = percent_rank(markov_score),
    combined_igm_score =
      0.55 * if_score_01 +
      0.35 * graph_score_01 +
      0.10 * markov_score_01
  ) %>%
  arrange(desc(combined_igm_score))

top_150_if_graph_markov_case10 <- case10_combo %>%
  slice_head(n = 150)

summary(top_150_if_graph_markov_case10$if_score)
summary(top_150_if_graph_markov_case10$graph_score)
summary(top_150_if_graph_markov_case10$markov_score)

writeLines(as.character(top_150_if_graph_markov_case10$transaction_id))

write.csv(
  top_150_if_graph_markov_case10,
  "~/Desktop/top_150_if_graph_markov_case10.csv",
  row.names = FALSE
)
```

