---
title: "Supermarket Case - EDA & Anomaly Exploration"
author: "Tristan & Hannes"
date: "2026-04-20"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Exploratory Data Analysis

## Libraries 

The analysis uses the tidyverse for data manipulation and visualization, knitr and kableExtra for table formatting in the report, and hexbin for density-based visualizations (hexagonal binning).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(hexbin)
```
 
## Raw Dataset

The dataset is read as raw text using readLines(), where each line represents a single transaction. A tibble is then created with a unique transaction_id for each row and a basket column containing the raw transaction string. The head() function is used to inspect the first observations.
 
```{r}
lines <- readLines("~/Desktop/supermarket case/supermarket.csv", warn = FALSE)

supermarket_raw <- tibble(
  transaction_id = 1:length(lines),
  basket = lines
)

head(supermarket_raw)
```

## Transfotmation into a structured format 

The raw transaction strings are transformed into a structured format. Each basket is split into individual product entries, and each entry is further separated into department, time (between scans), and price. These variables are then converted to numeric types. Finally, the data is inspected using head(), structure, summary statistics, and a check of unique transaction IDs.

```{r}
long_data <- supermarket_raw %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price)
  )

head(long_data)
str(long_data)
summary(long_data)
length(unique(long_data$transaction_id))
```

The dataset contains over 2.1 million item-level observations across 100,000 transactions.

The distribution of time between scans is right-skewed, with a median of 60 seconds but extreme values reaching up to 2,420 seconds, indicating the presence of long pauses.

Price distribution is also right-skewed, with most items priced around €2, but occasional high-value purchases.

Additionally, approximately 5% of the data contains missing values, which requires further data cleaning.

## Mapping

A mapping is created to translate numerical department codes into descriptive category names, then applied to the dataset, adding a new variable (department_name) that improves interpretability. The updated data is inspected using head().

```{r}
dept_map <- c(
  "1" = "Bakery & Pastry",
  "2" = "Beer & Wine",
  "3" = "Books & Magazines",
  "4" = "Candy & Chips",
  "5" = "Care & Hygiene",
  "6" = "Cereals & Spreads",
  "7" = "Cheese & Tapas",
  "8" = "Dairy & Eggs",
  "9" = "Freezer",
  "10" = "Fruit & Vegetables",
  "11" = "Household & Pet",
  "12" = "Meat & Fish",
  "13" = "Pasta & Rice",
  "14" = "Salads & Meals",
  "15" = "Sauces & Spices",
  "16" = "Soda & Juices",
  "17" = "Special Diet",
  "18" = "Vegetarian & Vegan"
)

long_data <- long_data %>%
  mutate(department_name = dept_map[as.character(department)])

head(long_data)
```

Numeric department codes were mapped to descriptive category names. This transformation allows for a more meaningful analysis of customer purchasing behavior, enabling comparisons across product categories and identification of common shopping patterns.

## Data Cleaning

The dataset is cleaned by removing observations with missing or invalid values (for example, NA, negative time or price). A new variable, price_per_second, is then created to capture the ratio of price to time between scans, serving as an additional behavioral feature.

```{r}
long_data_clean <- long_data %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  ) %>%
  mutate(
    price_per_second = ifelse(time > 0, price / time, NA_real_)
  )
```

## Descriptive Analysis of Basket Characteristics

The data is aggregated at the transaction level, turning item-level data into basket-level features. For each transaction, key metrics are calculated, including basket size (total_items), total and average price, total and average time between scans, variability measures, and price-related indicators. The number of unique departments is also included.

The resulting dataset is then checked using basic outputs such as head(), structure, and summary statistics.

```{r}
basket_data <- long_data_clean %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    max_time = max(time, na.rm = TRUE),
    min_time = min(time, na.rm = TRUE),
    price_per_second = ifelse(total_time > 0, total_price / total_time, NA_real_),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

head(basket_data)
str(basket_data)
summary(basket_data)
```

## Summary Statistic Table 

A summary statistics table is created for the main basket-level variables. The selected variables are reshaped into long format, grouped by metric, and summarized using mean, median, standard deviation, selected percentiles, and maximum values. The metric names are then relabeled for readability, numeric values are rounded, and the final table is formatted using kable and kableExtra for clear presentation in the report.

```{r}
stats_table <- basket_data %>%
  select(total_items, total_price, total_time, avg_time_per_item, sd_time_per_item) %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(
    Mean = mean(Value, na.rm = TRUE),
    Median = median(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),
    P25 = quantile(Value, 0.25, na.rm = TRUE),
    P75 = quantile(Value, 0.75, na.rm = TRUE),
    Max = max(Value, na.rm = TRUE)
  )

stats_table <- stats_table %>%
  mutate(Metric = recode(Metric,
    total_items = "Number of items",
    total_price = "Total price (€)",
    total_time = "Total scan time (sec)",
    avg_time_per_item = "Avg time per item (sec)",
    sd_time_per_item = "Scan time variability"
  ),
  across(-Metric, ~round(.x, 2))
  )

kable(stats_table, digits = 2, align = "lrrrrrr",
      caption = "Summary statistics of customer basket behavior", booktabs = TRUE) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, "Statistics" = 6))
```

The typical customer basket contains around 19-20 items, with total spending around €48.6. The distribution of total price is right-skewed, meaning there are some higher-value transactions.

Customers spend about 30 minutes per shopping session, with most trips lasting between 23 and 42 minutes.

The average time per item is relatively high (around 99 seconds), and there is quite a lot of variation within baskets. This suggests that customers do not scan items continuously, but take pauses as part of the shopping process.

Finally, the presence of extreme values in time-related variables points to outliers, such as very long pauses or possibly unclosed sessions.

## 3. EDA - Anomaly Exploration

### Tristan: Price-Based Anomalies

#### 1. Very Low / Very High Total Price

```{r tristan_price_distribution}
# Visualize the distribution to see outliers
ggplot(basket_data, aes(y = total_price)) +
  geom_boxplot(fill = "skyblue", outlier.color = "red", outlier.shape = 16) +
  theme_minimal() +
  labs(title = "Tristan: Distribution of Total Basket Price", 
       y = "Total Price", 
       subtitle = "Red dots indicate potential price anomalies")

# Isolate the extreme price baskets (e.g., bottom 1% and top 1%)
price_bounds <- quantile(basket_data$total_price, probs = c(0.01, 0.99), na.rm = TRUE)
extreme_price_baskets <- basket_data %>%
  filter(total_price <= price_bounds[1] | total_price >= price_bounds[2]) %>%
  arrange(desc(total_price))

head(extreme_price_baskets)
```

#### 2. Expensive Items in Small Baskets

```{r tristan_expensive_small}
# Scatter plot looking for high max_price isolated on the left side (low total_items)
ggplot(basket_data, aes(x = total_items, y = max_price)) +
  geom_point(alpha = 0.5, color = "darkred") +
  theme_minimal() +
  labs(title = "Tristan: Max Item Price vs. Basket Size", 
       x = "Total Items in Basket", 
       y = "Max Price of a Single Item")

# Filter for baskets with 3 or fewer items but containing an item in the top 5% most expensive
high_price_threshold <- quantile(basket_data$max_price, 0.95, na.rm = TRUE)
expensive_small_baskets <- basket_data %>%
  filter(total_items <= 3 & max_price >= high_price_threshold) %>%
  arrange(desc(max_price))

head(expensive_small_baskets)
```

### Hannes: Time-Based Anomalies

#### 1. Extremely Fast / Extremely Slow Scanning

```{r hannes_time_distribution}
# Visualize the distribution of average time per item
ggplot(basket_data, aes(x = avg_time_per_item)) +
  geom_histogram(bins = 60, fill = "lightgreen", color = "black") +
  theme_minimal() +
  labs(title = "Hannes: Distribution of Average Scan Time per Item", 
       x = "Average Time per Item (seconds)", 
       y = "Frequency")


ggplot(basket_data, aes(y = avg_time_per_item)) +
  geom_boxplot(fill = "lightgreen", outlier.color = "red", outlier.shape = 16) +
  theme_minimal() +
  labs(title = "Hannes: Outliers in Average Scan Time", 
       y = "Average Time per Item (seconds)")

# Isolate extreme time anomalies (e.g., bottom 1% and top 1% of scan speeds)
time_bounds <- quantile(basket_data$avg_time_per_item, probs = c(0.01, 0.99), na.rm = TRUE)
extreme_time_baskets <- basket_data %>%
  filter(avg_time_per_item <= time_bounds[1] | avg_time_per_item >= time_bounds[2]) %>%
  arrange(desc(avg_time_per_item))

head(extreme_time_baskets)
```

#### 2. Total Time vs Total Items Checkout

```{r hannes_total_time_items}
# Optional: Checking total_time vs total_items to see if some massive baskets were scanned impossibly fast
ggplot(basket_data, aes(x = total_items, y = total_time)) +
  geom_point(alpha = 0.4, color = "purple") +
  theme_minimal() +
  labs(title = "Hannes: Total Time vs Total Items",
       x = "Total Items",
       y = "Total Time (seconds)")
```
