---
title: "Case 1 - Behavioral / DTW Fraud Detection Prototype"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  reference_data_path: "supermarket.csv"
  test_data_path: null
  expected_fraud_share: 0.15
  output_rows_file: "predicted_fraud_rows.txt"
  seq_len: 25
  time_cap: 300
  price_cap: 20
  max_reference_rows: null
---

# Overview

This R Markdown file implements several **behavioral fraud-detection methods** for the supermarket self-checkout case. The focus is on the **sequence and timing behavior** of transactions rather than only on basket totals.

The workflow is designed so that you can:

- use the main `supermarket.csv` file as an unlabeled reference dataset,
- input a **new testing dataset** in the **same raw format**,
- score all rows in the testing dataset,
- return the **top suspected fraud cases**, and
- write the predicted row numbers to a text file with **one row number per line**.

The methods below do **not** require fraud labels in the training data. They learn what *normal* checkout behavior looks like from the reference data and then flag the most behaviorally unusual transactions in the testing data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Libraries

```{r libraries}
library(tidyverse)
```

# User Inputs

Set the file paths here before running the notebook.

- `reference_data_path`: the large unlabeled supermarket dataset used to learn normal behavior
- `test_data_path`: the new testing dataset you want to score
- `expected_fraud_share`: set to `0.15` because the testing set contains 15% fraud
- `output_rows_file`: file where the predicted suspicious row numbers will be written

```{r inputs}
reference_data_path <- params$reference_data_path
test_data_path      <- params$test_data_path
expected_fraud_share <- params$expected_fraud_share
output_rows_file     <- params$output_rows_file
seq_len              <- params$seq_len
time_cap             <- params$time_cap
price_cap            <- params$price_cap
max_reference_rows   <- params$max_reference_rows

if (is.null(test_data_path)) {
  message("No separate testing dataset was supplied. For development only, the reference file will also be used as the test file.")
  test_data_path <- reference_data_path
}
```

# Helper Functions

## 1. Reading and parsing raw transaction files

The raw data is expected to have the same structure as `supermarket.csv`: one transaction per row, where each row contains comma-separated item records in the form `department time price`.

```{r helper-read-parse}
read_transaction_file <- function(file_path, max_rows = NULL) {
  lines <- readLines(file_path, warn = FALSE)
  if (!is.null(max_rows)) {
    lines <- lines[seq_len(min(length(lines), max_rows))]
  }
  tibble(
    transaction_id = seq_along(lines),
    basket = lines
  )
}

parse_transaction_file <- function(raw_tbl) {
  raw_tbl %>%
    separate_rows(basket, sep = ",") %>%
    group_by(transaction_id) %>%
    mutate(item_order = row_number()) %>%
    ungroup() %>%
    mutate(basket = str_trim(basket)) %>%
    separate(
      basket,
      into = c("department", "time", "price"),
      sep = "\\s+",
      fill = "right",
      extra = "drop"
    ) %>%
    mutate(
      department = as.integer(department),
      time = as.numeric(time),
      price = as.numeric(price)
    )
}

clean_long_data <- function(long_data) {
  long_data %>%
    filter(
      !is.na(transaction_id),
      !is.na(item_order),
      !is.na(department),
      !is.na(time),
      !is.na(price),
      department >= 1,
      department <= 18,
      time >= 0,
      price >= 0
    )
}
```

## 2. Behavioral feature engineering

These features summarize each basket from a **behavioral** point of view.

```{r helper-features}
build_behavior_features <- function(long_data) {
  seq_base <- long_data %>%
    arrange(transaction_id, item_order) %>%
    group_by(transaction_id) %>%
    mutate(
      next_department = lead(department),
      same_department_as_previous = department == lag(department),
      fast_scan = time <= 10,
      long_pause = time >= 180,
      zero_time = time == 0
    ) %>%
    summarise(
      total_items = n(),
      total_price = sum(price, na.rm = TRUE),
      total_time = sum(time, na.rm = TRUE),
      avg_time_per_item = mean(time, na.rm = TRUE),
      median_time_per_item = median(time, na.rm = TRUE),
      sd_time_per_item = sd(time, na.rm = TRUE),
      min_time = min(time, na.rm = TRUE),
      max_time = max(time, na.rm = TRUE),
      avg_price_per_item = mean(price, na.rm = TRUE),
      sd_price = sd(price, na.rm = TRUE),
      unique_departments = n_distinct(department),
      fast_scan_share = mean(fast_scan, na.rm = TRUE),
      long_pause_share = mean(long_pause, na.rm = TRUE),
      zero_time_share = mean(zero_time, na.rm = TRUE),
      same_department_share = mean(replace_na(same_department_as_previous, FALSE)),
      department_switch_rate = mean(department != lag(department), na.rm = TRUE),
      price_time_correlation = suppressWarnings(cor(price, time, use = "pairwise.complete.obs")),
      .groups = "drop"
    ) %>%
    mutate(
      sd_time_per_item = replace_na(sd_time_per_item, 0),
      sd_price = replace_na(sd_price, 0),
      price_time_correlation = replace_na(price_time_correlation, 0),
      burstiness = ifelse(avg_time_per_item > 0, sd_time_per_item / avg_time_per_item, 0),
      items_per_department = ifelse(unique_departments > 0, total_items / unique_departments, 0),
      price_per_second = ifelse(total_time > 0, total_price / total_time, 0)
    )

  seq_base
}
```

## 3. Sequence construction for DTW

For the DTW-based methods, each transaction is converted into a multivariate sequence using:

- time between scans,
- item price, and
- department code.

To make sequences more comparable, each dimension is capped and scaled.

```{r helper-sequences}
rescale_sequence <- function(df_txn, seq_len = 25, time_cap = 300, price_cap = 20) {
  df_txn <- df_txn %>% arrange(item_order)

  x_old <- seq_len(nrow(df_txn))
  x_new <- seq(1, nrow(df_txn), length.out = seq_len)

  time_scaled <- pmin(df_txn$time, time_cap) / time_cap
  price_scaled <- pmin(df_txn$price, price_cap) / price_cap
  dept_scaled <- (df_txn$department - 1) / 17

  interp_time <- approx(x_old, time_scaled, xout = x_new, method = "linear", rule = 2)$y
  interp_price <- approx(x_old, price_scaled, xout = x_new, method = "linear", rule = 2)$y
  interp_dept <- approx(x_old, dept_scaled, xout = x_new, method = "linear", rule = 2)$y

  cbind(interp_time, interp_price, interp_dept)
}

build_sequence_list <- function(long_data, seq_len = 25, time_cap = 300, price_cap = 20) {
  split(long_data, long_data$transaction_id) %>%
    purrr::map(~ rescale_sequence(.x, seq_len = seq_len, time_cap = time_cap, price_cap = price_cap))
}
```

## 4. Custom multivariate DTW distance

To keep the notebook portable, DTW is implemented directly in base R. The local step cost is the Euclidean distance between the resampled feature vectors.

```{r helper-dtw}
dtw_distance_multivariate <- function(seq_a, seq_b) {
  n <- nrow(seq_a)
  m <- nrow(seq_b)

  cost <- matrix(Inf, nrow = n + 1, ncol = m + 1)
  cost[1, 1] <- 0

  for (i in seq_len(n)) {
    for (j in seq_len(m)) {
      local_cost <- sqrt(sum((seq_a[i, ] - seq_b[j, ])^2))
      cost[i + 1, j + 1] <- local_cost + min(
        cost[i, j + 1],
        cost[i + 1, j],
        cost[i, j]
      )
    }
  }

  cost[n + 1, m + 1] / (n + m)
}
```

## 5. Prototype selection for DTW

Instead of comparing every test transaction with every reference transaction, the code builds a small set of **behavior prototypes** from the reference data. These prototypes are selected within basket-size bins so that comparisons stay behaviorally meaningful.

```{r helper-prototypes}
make_size_bin <- function(total_items) {
  case_when(
    total_items <= 10 ~ "small",
    total_items <= 20 ~ "medium",
    total_items <= 35 ~ "large",
    TRUE ~ "very_large"
  )
}

select_behavior_prototypes <- function(feature_tbl, n_clusters = 3, max_per_bin = 3000) {
  selected <- list()

  for (bin in unique(feature_tbl$size_bin)) {
    df_bin <- feature_tbl %>% filter(size_bin == bin)
    if (nrow(df_bin) == 0) next

    if (!is.null(max_per_bin) && nrow(df_bin) > max_per_bin) {
      set.seed(123)
      df_bin <- df_bin %>% slice_sample(n = max_per_bin)
    }

    mat <- df_bin %>%
      select(avg_time_per_item, sd_time_per_item, fast_scan_share, long_pause_share,
             same_department_share, department_switch_rate, avg_price_per_item,
             unique_departments, burstiness) %>%
      scale()

    k <- min(n_clusters, nrow(df_bin))

    if (k == 1) {
      selected[[bin]] <- df_bin %>% slice(1) %>% pull(transaction_id)
    } else {
      km <- kmeans(mat, centers = k, nstart = 10)
      df_bin$cluster <- km$cluster

      medoids <- df_bin %>%
        group_by(cluster) %>%
        group_modify(~ {
          center <- km$centers[unique(.x$cluster), , drop = FALSE]
          xmat <- .x %>%
            select(avg_time_per_item, sd_time_per_item, fast_scan_share, long_pause_share,
                   same_department_share, department_switch_rate, avg_price_per_item,
                   unique_departments, burstiness) %>%
            scale(center = attr(mat, "scaled:center"), scale = attr(mat, "scaled:scale"))
          dists <- rowSums((as.matrix(xmat) - matrix(center, nrow = nrow(xmat), ncol = ncol(xmat), byrow = TRUE))^2)
          .x[which.min(dists), , drop = FALSE]
        }) %>%
        ungroup() %>%
        pull(transaction_id)

      selected[[bin]] <- medoids
    }
  }

  enframe(selected, name = "size_bin", value = "prototype_ids")
}
```

## 6. Transition-likelihood model

This second behavioral approach does not use DTW. Instead, it estimates how *unlikely* the observed department-to-department transitions are under normal shopping behavior.

```{r helper-transition-model}
fit_transition_model <- function(long_data) {
  transition_counts <- long_data %>%
    arrange(transaction_id, item_order) %>%
    group_by(transaction_id) %>%
    mutate(next_department = lead(department)) %>%
    ungroup() %>%
    filter(!is.na(next_department)) %>%
    count(department, next_department, name = "n")

  all_pairs <- expand_grid(department = 1:18, next_department = 1:18)

  transition_probs <- all_pairs %>%
    left_join(transition_counts, by = c("department", "next_department")) %>%
    mutate(n = replace_na(n, 0)) %>%
    group_by(department) %>%
    mutate(prob = (n + 1) / (sum(n) + 18)) %>%
    ungroup()

  dept_time_stats <- long_data %>%
    group_by(department) %>%
    summarise(
      med_time = median(time, na.rm = TRUE),
      mad_time = mad(time, constant = 1, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(mad_time = ifelse(mad_time <= 0, 1, mad_time))

  list(
    transition_probs = transition_probs,
    dept_time_stats = dept_time_stats
  )
}

score_transition_anomaly <- function(long_data, model) {
  transition_scores <- long_data %>%
    arrange(transaction_id, item_order) %>%
    group_by(transaction_id) %>%
    mutate(next_department = lead(department)) %>%
    ungroup() %>%
    filter(!is.na(next_department)) %>%
    left_join(model$transition_probs, by = c("department", "next_department")) %>%
    mutate(step_nll = -log(prob)) %>%
    group_by(transaction_id) %>%
    summarise(
      transition_nll = mean(step_nll, na.rm = TRUE),
      .groups = "drop"
    )

  time_scores <- long_data %>%
    left_join(model$dept_time_stats, by = "department") %>%
    mutate(time_robust_z = abs(time - med_time) / mad_time) %>%
    group_by(transaction_id) %>%
    summarise(
      time_behavior_score = mean(time_robust_z, na.rm = TRUE),
      .groups = "drop"
    )

  transition_scores %>%
    full_join(time_scores, by = "transaction_id") %>%
    mutate(
      transition_nll = replace_na(transition_nll, 0),
      time_behavior_score = replace_na(time_behavior_score, 0),
      transition_behavior_score = transition_nll + time_behavior_score
    )
}
```

## 7. Multivariate behavioral anomaly score

The third approach summarizes each transaction into behavioral features and scores it using **Mahalanobis distance** relative to the reference data.

```{r helper-mahalanobis}
fit_behavior_baseline <- function(feature_tbl) {
  baseline_vars <- c(
    "total_items", "total_time", "avg_time_per_item", "sd_time_per_item",
    "fast_scan_share", "long_pause_share", "zero_time_share",
    "same_department_share", "department_switch_rate", "burstiness",
    "avg_price_per_item", "unique_departments", "price_per_second"
  )

  train_mat <- feature_tbl %>% select(all_of(baseline_vars)) %>% as.matrix()

  list(
    baseline_vars = baseline_vars,
    center = colMeans(train_mat, na.rm = TRUE),
    cov = cov(train_mat, use = "pairwise.complete.obs")
  )
}

score_mahalanobis_behavior <- function(feature_tbl, model) {
  x <- feature_tbl %>% select(all_of(model$baseline_vars)) %>% as.matrix()

  scores <- mahalanobis(
    x = x,
    center = model$center,
    cov = model$cov
  )

  tibble(
    transaction_id = feature_tbl$transaction_id,
    mahalanobis_behavior_score = scores
  )
}
```

## 8. Rank-based ensemble

Because each method captures a different aspect of suspicious behavior, the final ranking combines them in a simple ensemble.

```{r helper-ensemble}
rank_normalize <- function(x) {
  if (all(is.na(x))) return(rep(NA_real_, length(x)))
  (rank(x, ties.method = "average", na.last = "keep") - 1) / (sum(!is.na(x)) - 1)
}

make_final_ensemble <- function(score_tbl) {
  score_tbl %>%
    mutate(
      dtw_rank = rank_normalize(dtw_score),
      transition_rank = rank_normalize(transition_behavior_score),
      mahalanobis_rank = rank_normalize(mahalanobis_behavior_score),
      ensemble_score = rowMeans(
        cbind(dtw_rank, transition_rank, mahalanobis_rank),
        na.rm = TRUE
      )
    ) %>%
    arrange(desc(ensemble_score))
}
```

# Load Reference and Testing Data

The reference file is used to learn normal behavior. The testing file is then scored using the methods above.

```{r load-data}
reference_raw <- reference_data_path <- "/Users/macbook/Desktop/supermarket case/supermarket.csv"
test_raw <- test_data_path <- "/Users/macbook/Desktop/case12.csv"

read_transaction_file <- function(file_path, max_rows = NULL) {
  lines <- readLines(file_path, warn = FALSE)
  if (!is.null(max_rows)) {
    lines <- lines[seq_len(min(length(lines), max_rows))]
  }
  tibble(
    transaction_id = seq_along(lines),
    basket = lines
  )
}

reference_data_path <- "/Users/macbook/Desktop/supermarket case/supermarket.csv"

reference_raw <- read_transaction_file(reference_data_path, max_rows = max_reference_rows)

class(reference_raw)
head(reference_raw)

reference_data_path <- "/Users/macbook/Desktop/supermarket case/supermarket.csv"
test_data_path <- "/Users/macbook/Desktop/supermarket case/supermarket.csv"

reference_raw <- read_transaction_file(reference_data_path, max_rows = max_reference_rows)
test_raw <- read_transaction_file(test_data_path)

reference_long <- reference_raw %>% parse_transaction_file() %>% clean_long_data()
test_long <- test_raw %>% parse_transaction_file() %>% clean_long_data()

cat("Reference transactions:", n_distinct(reference_long$transaction_id), "\n")
cat("Test transactions:", n_distinct(test_long$transaction_id), "\n")
```

# Build Features and Sequences

```{r build-features-sequences}
reference_features <- build_behavior_features(reference_long) %>%
  mutate(size_bin = make_size_bin(total_items))

test_features <- build_behavior_features(test_long) %>%
  mutate(size_bin = make_size_bin(total_items))

reference_sequences <- build_sequence_list(reference_long, seq_len = seq_len, time_cap = time_cap, price_cap = price_cap)
test_sequences <- build_sequence_list(test_long, seq_len = seq_len, time_cap = time_cap, price_cap = price_cap)
```

# Method 1: DTW Distance to Normal Behavior Prototypes

This method compares the testing transactions to a small set of **normal behavior prototypes** learned from the reference data. Transactions with larger DTW distances are treated as more suspicious.

```{r method-1-dtw}
prototype_tbl <- select_behavior_prototypes(reference_features, n_clusters = 3, max_per_bin = 3000)
prototype_ids <- unlist(prototype_tbl$prototype_ids)

prototype_lookup <- reference_features %>%
  filter(transaction_id %in% prototype_ids) %>%
  select(transaction_id, size_bin)

score_dtw_against_prototypes <- function(test_features, test_sequences, prototype_lookup, reference_sequences) {
  results <- vector("list", nrow(test_features))

  for (i in seq_len(nrow(test_features))) {
    tx_id <- test_features$transaction_id[i]
    tx_bin <- test_features$size_bin[i]
    tx_seq <- test_sequences[[as.character(tx_id)]]

    candidate_prototypes <- prototype_lookup %>% filter(size_bin == tx_bin)
    if (nrow(candidate_prototypes) == 0) {
      candidate_prototypes <- prototype_lookup
    }

    dists <- purrr::map_dbl(candidate_prototypes$transaction_id, function(pid) {
      ref_seq <- reference_sequences[[as.character(pid)]]
      dtw_distance_multivariate(tx_seq, ref_seq)
    })

    results[[i]] <- tibble(
      transaction_id = tx_id,
      dtw_score = min(dists, na.rm = TRUE),
      dtw_mean_score = mean(dists, na.rm = TRUE)
    )
  }

  bind_rows(results)
}

dtw_scores <- score_dtw_against_prototypes(
  test_features = test_features,
  test_sequences = test_sequences,
  prototype_lookup = prototype_lookup,
  reference_sequences = reference_sequences
)

head(dtw_scores %>% arrange(desc(dtw_score)), 10)
```

# Method 2: Transition-Likelihood / Behavioral Sequence Anomaly

This method scores each transaction based on:

- how rare its department transitions are, and
- how unusual its scan times are relative to normal timings for each department.

```{r method-2-transition}
transition_model <- fit_transition_model(reference_long)
transition_scores <- score_transition_anomaly(test_long, transition_model)

head(transition_scores %>% arrange(desc(transition_behavior_score)), 10)
```

# Method 3: Multivariate Behavioral Anomaly Score

This method uses basket-level behavioral summaries and measures how far each testing transaction is from the normal behavioral baseline.

```{r method-3-mahalanobis}
behavior_model <- fit_behavior_baseline(reference_features)
mahalanobis_scores <- score_mahalanobis_behavior(test_features, behavior_model)

head(mahalanobis_scores %>% arrange(desc(mahalanobis_behavior_score)), 10)
```

# Combine the Methods into a Final Fraud Ranking

The three methods are merged and combined into one rank-based ensemble score.

```{r combine-methods}
all_scores <- test_features %>%
  select(transaction_id, total_items, total_price, total_time, avg_time_per_item, size_bin) %>%
  left_join(dtw_scores, by = "transaction_id") %>%
  left_join(transition_scores, by = "transaction_id") %>%
  left_join(mahalanobis_scores, by = "transaction_id")

final_scores <- make_final_ensemble(all_scores)

head(final_scores %>% select(
  transaction_id, total_items, total_price, total_time,
  dtw_score, transition_behavior_score,
  mahalanobis_behavior_score, ensemble_score
), 20)
```

# Select the Top Suspected Fraud Cases

The testing datasets are assumed to contain **1000 rows** and **15% fraud**. Therefore, the code returns the **150 most suspicious rows** by default. The code is written more generally, so it will always return `round(0.15 * number_of_test_rows)` transactions.

```{r select-top-cases}
n_test <- nrow(test_raw)
expected_fraud_n <- round(expected_fraud_share * n_test)

predicted_fraud_rows <- final_scores %>%
  slice_max(order_by = ensemble_score, n = expected_fraud_n, with_ties = FALSE) %>%
  arrange(transaction_id) %>%
  pull(transaction_id)

cat("Number of predicted fraud cases returned:", length(predicted_fraud_rows), "\n")
cat("First predicted row numbers:\n")
print(head(predicted_fraud_rows, 20))
```

# Output: Row Numbers Only (One Per Line)

This is the output format you asked for. The predicted row numbers are written to a text file and also printed line by line.

```{r output-rows}
writeLines(as.character(predicted_fraud_rows), con = output_rows_file)

cat("Predicted fraud row numbers written to:", output_rows_file, "\n\n")
cat("Predicted fraud row numbers (one per line):\n")
cat(predicted_fraud_rows, sep = "\n")
```

# Optional: Inspect the Top-Ranked Transactions in More Detail

This section helps you understand *why* a transaction was ranked highly.

```{r inspect-top-cases}
inspection_tbl <- final_scores %>%
  filter(transaction_id %in% predicted_fraud_rows) %>%
  arrange(desc(ensemble_score)) %>%
  select(
    transaction_id, size_bin, total_items, total_price, total_time,
    avg_time_per_item, dtw_score, transition_behavior_score,
    mahalanobis_behavior_score, ensemble_score
  )

head(inspection_tbl, 25)
```

# Suggested Interpretation

The behavioral logic behind the three methods is slightly different:

1. **DTW method**: flags transactions whose scan sequence looks unlike normal checkout behavior.
2. **Transition-likelihood method**: flags transactions with unusual department orderings and unusual timing patterns.
3. **Mahalanobis method**: flags transactions with unusual overall behavioral summaries.

The final ensemble is useful because fraud may not always appear in exactly the same way. Some fraudulent baskets may have strange timing, others may have odd sequences, and others may simply look behaviorally inconsistent overall.

# How to Use This with a New Testing Dataset

To test a new dataset, change only these inputs near the start of the notebook:

```r
reference_data_path <- "path/to/supermarket.csv"
test_data_path <- "path/to/testing_dataset.csv"
expected_fraud_share <- 0.15
output_rows_file <- "predicted_fraud_rows.txt"
```

Then run the notebook again.

# Important Note

Because the reference dataset is unlabeled, these methods learn **normal behavior**, not confirmed fraud behavior. That means the output should be interpreted as:

- **most suspicious / most behaviorally unusual transactions**,
- not guaranteed fraud labels.

If you later get access to actual fraud labels for the testing data, you can evaluate performance using precision, recall, and ranking quality.
