---
title: "Case 1 - Improved Clustering-Based Fraud Detection Prototype"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  supermarket_reference_path: "supermarket.csv"
  cashier_data_path: "cashier_data.csv"
  test_data_path: null
  expected_fraud_share: 0.15
  output_rows_file: "predicted_fraud_ids.txt"
  k_kmeans_cashier: 18
  k_kmeans_combined: 16
  k_kmeans_selfcheckout: 18
  k_segmented_per_bin: 6
  max_rows_supermarket_for_combined: 20000
  max_rows_cashier_for_combined: 10000
  max_rows_supermarket_for_self: 25000
  seed: 123
  shrinkage_path: "shrinkage.csv"
---

# Overview

This R Markdown file implements an **improved clustering-based fraud detection workflow** for Case 1.

Compared with the earlier version, this file improves the method in four important ways:

1. **Cashier data is treated differently from self-checkout data**  
   Cashier baskets do not contain time, so cashier-based clustering uses only **shared non-time features**.

2. **Basket size is handled more carefully**  
   One method uses **size-segmented cashier clustering**, so baskets are compared with normal baskets of similar size.

3. **Self-checkout timing is still used where it makes sense**  
   A separate supermarket-only clustering method uses **time and behavioral features**, but only on self-checkout data.

4. **The final score is a weighted ensemble**  
   Instead of relying on one clustering result, the file combines several clustering-based anomaly signals.

The workflow is designed so that you can:

- use `cashier_data.csv` as a **clean no-fraud baseline**,
- use `supermarket.csv` as a **self-checkout reference**,
- input a new testing dataset in the same raw format,
- score all rows in the testing dataset,
- return the top **15% most suspicious transaction IDs**, and
- write those transaction IDs to a text file with **one ID per line**.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Libraries

```{r libraries}
library(tidyverse)
library(cluster)
setwd("/Users/tristankluth/Desktop/Uni Maastricht/Year 2/Period 5/eLab Business Case II/Case 1")
```

# User Inputs

```{r inputs}
set.seed(123)

supermarket_reference_path <- "/Users/tristankluth/Desktop/Uni Maastricht/Year 2/Period 5/eLab Business Case II/Case 1/supermarket.csv"
cashier_data_path          <- "/Users/tristankluth/Desktop/Uni Maastricht/Year 2/Period 5/eLab Business Case II/Case 1/cashier_data.csv"
test_data_path             <- "/Users/tristankluth/Desktop/Uni Maastricht/Year 2/Period 5/eLab Business Case II/Case 1/case97.csv"
expected_fraud_share       <- 0.15
output_rows_file           <- "predicted_fraud_rows.txt"
shrinkage_path             <- "/Users/tristankluth/Desktop/Uni Maastricht/Year 2/Period 5/eLab Business Case II/Case 1/shrinkage.csv"

k_kmeans_cashier           <- 12
k_pam_cashier              <- 12
k_kmeans_combined          <- 14
max_rows_pam_train         <- 5000
max_rows_supermarket_for_combined <- 15000
max_rows_cashier_for_combined     <- 15000
k_segmented_per_bin <- 6
max_rows_supermarket_for_self <- 20000
k_kmeans_selfcheckout <- 12

if (is.null(test_data_path)) {
  message("No separate testing dataset supplied. For development only, supermarket_reference_path will also be used as test data.")
  test_data_path <- supermarket_reference_path
}
```

# Helper Functions

## Read raw transaction files

```{r helper-read}
read_transaction_file <- function(file_path, source_name = "unknown") {
  first_line <- readLines(file_path, n = 1, warn = FALSE)

  if (grepl("^\\s*transaction_id\\s*,\\s*basket\\s*$", first_line, ignore.case = TRUE)) {
    df <- read.csv(file_path, stringsAsFactors = FALSE)

    embedded_match <- stringr::str_match(
      df$basket,
      "^\\s*Transaction\\s+([0-9]+)\\s*:\\s*(.*)$"
    )

    has_embedded_id <- !is.na(embedded_match[, 2])

    transaction_ids <- ifelse(
      has_embedded_id,
      as.integer(embedded_match[, 2]),
      if ("transaction_id" %in% names(df)) as.integer(df$transaction_id) else seq_len(nrow(df))
    )

    basket_text <- ifelse(
      has_embedded_id,
      embedded_match[, 3],
      df$basket
    )

    return(
      tibble(
        transaction_id = transaction_ids,
        basket = basket_text,
        source = source_name
      )
    )
  }

  lines <- readLines(file_path, warn = FALSE)

  embedded_match <- stringr::str_match(
    lines,
    "^\\s*Transaction\\s+([0-9]+)\\s*:\\s*(.*)$"
  )

  has_embedded_id <- !is.na(embedded_match[, 2])

  transaction_ids <- ifelse(
    has_embedded_id,
    as.integer(embedded_match[, 2]),
    seq_along(lines)
  )

  basket_text <- ifelse(
    has_embedded_id,
    embedded_match[, 3],
    lines
  )

  tibble(
    transaction_id = transaction_ids,
    basket = basket_text,
    source = source_name
  )
}
```

## Parse basket items

- `supermarket` and `test` items have the format: `department time price`
- `cashier` items have the format: `department price`

```{r helper-parse}
parse_transaction_file <- function(transaction_tbl) {
  transaction_tbl %>%
    tidyr::separate_rows(basket, sep = ",") %>%
    dplyr::mutate(
      basket = stringr::str_trim(basket)
    ) %>%
    dplyr::filter(basket != "") %>%
    dplyr::group_by(transaction_id) %>%
    dplyr::mutate(
      item_order = dplyr::row_number()
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      n_parts = stringr::str_count(basket, "\\S+")
    ) %>%
    tidyr::extract(
      basket,
      into = c("v1", "v2", "v3"),
      regex = "^\\s*(\\S+)\\s+(\\S+)(?:\\s+(\\S+))?\\s*$",
      remove = FALSE
    ) %>%
    dplyr::mutate(
      department = as.integer(v1),
      time = dplyr::case_when(
        source == "cashier" & n_parts == 2 ~ 0,
        n_parts == 3 ~ as.numeric(v2),
        TRUE ~ NA_real_
      ),
      price = dplyr::case_when(
        source == "cashier" & n_parts == 2 ~ as.numeric(v2),
        n_parts == 3 ~ as.numeric(v3),
        TRUE ~ NA_real_
      )
    ) %>%
    dplyr::select(transaction_id, department, time, price, source, item_order)
}

clean_long_data <- function(long_data) {
  long_data %>%
    dplyr::filter(
      !is.na(department),
      !is.na(price),
      !is.na(time),
      department >= 1,
      department <= 18,
      price >= 0,
      time >= 0
    )
}


prepare_shrinkage_lookup <- function(shrinkage_path) {
  shrinkage <- read.csv(shrinkage_path, stringsAsFactors = FALSE) %>%
    mutate(
      item_loss_rate = ifelse(items_expected > 0, items_lost / items_expected, 0),
      revenue_loss_rate = ifelse(revenue_expected > 0, revenue_lost / revenue_expected, 0),
      combined_shrinkage_score_raw = 0.4 * item_loss_rate + 0.6 * revenue_loss_rate
    )

  score_center <- median(shrinkage$combined_shrinkage_score_raw, na.rm = TRUE)
  score_scale <- IQR(shrinkage$combined_shrinkage_score_raw, na.rm = TRUE)
  if (is.na(score_scale) || score_scale == 0) score_scale <- 1

  shrinkage %>%
    transmute(
      department = as.integer(department),
      item_loss_rate,
      revenue_loss_rate,
      combined_shrinkage_score = (combined_shrinkage_score_raw - score_center) / score_scale
    )
}

attach_shrinkage_features <- function(long_data, shrinkage_lookup) {
  risk_cutoff <- quantile(shrinkage_lookup$combined_shrinkage_score, 0.75, na.rm = TRUE)

  long_data %>%
    left_join(shrinkage_lookup, by = "department") %>%
    mutate(
      item_loss_rate = replace_na(item_loss_rate, 0),
      revenue_loss_rate = replace_na(revenue_loss_rate, 0),
      combined_shrinkage_score = replace_na(combined_shrinkage_score, 0),
      high_risk_department_flag = combined_shrinkage_score >= risk_cutoff
    )
}

```

## Feature engineering

```{r helper-features}
build_basket_features <- function(long_data) {
  dept_shares <- long_data %>%
    count(transaction_id, source, department, name = "dept_count") %>%
    group_by(transaction_id, source) %>%
    mutate(dept_share = dept_count / sum(dept_count)) %>%
    ungroup() %>%
    mutate(department = paste0("dept_share_", department)) %>%
    select(transaction_id, source, department, dept_share) %>%
    pivot_wider(names_from = department, values_from = dept_share, values_fill = 0)

  base_features <- long_data %>%
    group_by(transaction_id, source) %>%
    summarise(
      total_items = n(),
      total_price = sum(price, na.rm = TRUE),
      avg_price = mean(price, na.rm = TRUE),
      median_price = median(price, na.rm = TRUE),
      sd_price = sd(price, na.rm = TRUE),
      max_price = max(price, na.rm = TRUE),
      min_price = min(price, na.rm = TRUE),
      total_time = sum(time, na.rm = TRUE),
      avg_time_per_item = mean(time, na.rm = TRUE),
      median_time_per_item = median(time, na.rm = TRUE),
      sd_time_per_item = sd(time, na.rm = TRUE),
      max_time = max(time, na.rm = TRUE),
      min_time = min(time, na.rm = TRUE),
      unique_departments = n_distinct(department),
      fast_scan_share = mean(time <= 10, na.rm = TRUE),
      long_pause_share = mean(time >= 180, na.rm = TRUE),
      zero_time_share = mean(time == 0, na.rm = TRUE),
      repeated_department_share = mean(department == lag(department), na.rm = TRUE),
      avg_department_risk = mean(combined_shrinkage_score, na.rm = TRUE),
      max_department_risk = max(combined_shrinkage_score, na.rm = TRUE),
      total_department_risk = sum(combined_shrinkage_score, na.rm = TRUE),
      avg_item_loss_rate = mean(item_loss_rate, na.rm = TRUE),
      avg_revenue_loss_rate = mean(revenue_loss_rate, na.rm = TRUE),
      high_risk_item_share = mean(high_risk_department_flag, na.rm = TRUE),
      high_risk_price_share = ifelse(sum(price, na.rm = TRUE) > 0,
                                     sum(price * high_risk_department_flag, na.rm = TRUE) / sum(price, na.rm = TRUE),
                                     0),
      risk_weighted_price = sum(price * pmax(combined_shrinkage_score, 0), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      sd_price = replace_na(sd_price, 0),
      sd_time_per_item = replace_na(sd_time_per_item, 0),
      repeated_department_share = replace_na(repeated_department_share, 0),
      avg_department_risk = replace_na(avg_department_risk, 0),
      max_department_risk = replace_na(max_department_risk, 0),
      total_department_risk = replace_na(total_department_risk, 0),
      avg_item_loss_rate = replace_na(avg_item_loss_rate, 0),
      avg_revenue_loss_rate = replace_na(avg_revenue_loss_rate, 0),
      high_risk_item_share = replace_na(high_risk_item_share, 0),
      high_risk_price_share = replace_na(high_risk_price_share, 0),
      risk_weighted_price = replace_na(risk_weighted_price, 0),
      avg_price_per_item = ifelse(total_items > 0, total_price / total_items, NA_real_),
      price_per_second = ifelse(total_time > 0, total_price / total_time, 0),
      items_per_department = ifelse(unique_departments > 0, total_items / unique_departments, NA_real_),
      max_price_share = ifelse(total_price > 0, max_price / total_price, NA_real_)
    )

  base_features %>%
    left_join(dept_shares, by = c("transaction_id", "source"))
}

add_size_bin <- function(df) {
  df %>%
    mutate(
      size_bin = case_when(
        total_items <= 10 ~ "small",
        total_items <= 20 ~ "medium",
        total_items <= 35 ~ "large",
        TRUE ~ "very_large"
      )
    )
}
```

## Residual feature from cashier baseline

```{r helper-residual}
add_residual_feature <- function(cashier_df, other_df_list) {
  cashier_model_data <- cashier_df %>%
    filter(!is.na(total_price), !is.na(total_items))

  price_model_cashier <- lm(total_price ~ total_items, data = cashier_model_data)

  add_one <- function(df) {
    df %>%
      mutate(
        expected_total_price_cashier = predict(price_model_cashier, newdata = df),
        residual_total_price_cashier = total_price - expected_total_price_cashier
      )
  }

  list(
    cashier = add_one(cashier_df),
    others = purrr::map(other_df_list, add_one)
  )
}
```

## Alignment and scaling helpers

```{r helper-align-scale}
align_feature_columns <- function(reference_df, new_df, feature_names) {
  missing_in_reference <- setdiff(feature_names, names(reference_df))
  missing_in_new <- setdiff(feature_names, names(new_df))

  if (length(missing_in_reference) > 0) {
    reference_df[missing_in_reference] <- 0
  }

  if (length(missing_in_new) > 0) {
    new_df[missing_in_new] <- 0
  }

  list(
    reference = reference_df %>% select(all_of(feature_names)),
    new = new_df %>% select(all_of(feature_names))
  )
}

fit_robust_scaler <- function(df) {
  med <- df %>% summarise(across(everything(), ~ median(.x, na.rm = TRUE)))
  iqr <- df %>% summarise(across(everything(), ~ IQR(.x, na.rm = TRUE)))
  iqr[iqr == 0 | is.na(iqr)] <- 1
  list(median = med, iqr = iqr, feature_names = names(df))
}

apply_robust_scaler <- function(df, scaler) {
  x <- as.matrix(df[, scaler$feature_names, drop = FALSE])
  x <- sweep(x, 2, as.numeric(scaler$median[1, ]), FUN = "-")
  x <- sweep(x, 2, as.numeric(scaler$iqr[1, ]), FUN = "/")
  x[is.na(x)] <- 0
  x
}
```

## Distance and scoring helpers

```{r helper-distance}
row_distance_to_centers <- function(x_mat, centers_mat) {
  n <- nrow(x_mat)
  k <- nrow(centers_mat)
  out <- matrix(NA_real_, nrow = n, ncol = k)

  for (j in seq_len(k)) {
    diffs <- sweep(x_mat, 2, centers_mat[j, ], FUN = "-")
    out[, j] <- sqrt(rowSums(diffs^2))
  }

  out
}

safe_rank_score <- function(x) {
  rank(x, ties.method = "average") / length(x)
}
```

# Load and Prepare the Datasets

```{r load-data}
supermarket_raw <- read_transaction_file(supermarket_reference_path, source_name = "supermarket")
cashier_raw     <- read_transaction_file(cashier_data_path, source_name = "cashier")
test_raw        <- read_transaction_file(test_data_path, source_name = "test")

cat("Rows in supermarket reference:", nrow(supermarket_raw), "\n")
cat("Rows in cashier data:", nrow(cashier_raw), "\n")
cat("Rows in test data:", nrow(test_raw), "\n")

shrinkage_lookup <- prepare_shrinkage_lookup(shrinkage_path)
```

```{r parse-clean}
supermarket_long <- supermarket_raw %>% parse_transaction_file() %>% clean_long_data() %>% attach_shrinkage_features(shrinkage_lookup)
cashier_long     <- cashier_raw %>% parse_transaction_file() %>% clean_long_data() %>% attach_shrinkage_features(shrinkage_lookup)
test_long        <- test_raw %>% parse_transaction_file() %>% clean_long_data() %>% attach_shrinkage_features(shrinkage_lookup)

cat("Rows in supermarket_long:", nrow(supermarket_long), "\n")
cat("Rows in cashier_long:", nrow(cashier_long), "\n")
cat("Rows in test_long:", nrow(test_long), "\n")
```

```{r build-features}
supermarket_features <- supermarket_long %>% build_basket_features() %>% add_size_bin()
cashier_features     <- cashier_long %>% build_basket_features() %>% add_size_bin()
test_features        <- test_long %>% build_basket_features() %>% add_size_bin()

residual_objects <- add_residual_feature(
  cashier_df = cashier_features,
  other_df_list = list(supermarket = supermarket_features, test = test_features)
)

cashier_features     <- residual_objects$cashier
supermarket_features <- residual_objects$others$supermarket
test_features        <- residual_objects$others$test

cat("Supermarket baskets after cleaning:", nrow(supermarket_features), "\n")
cat("Cashier baskets after cleaning:", nrow(cashier_features), "\n")
cat("Test baskets after cleaning:", nrow(test_features), "\n")
```

# Feature Sets

## Shared features
These are used for cashier-based methods and exclude time variables.

```{r feature-sets}
dept_feature_names <- paste0("dept_share_", 1:18)

shared_feature_names <- c(
  "total_items", "total_price", "avg_price", "median_price", "sd_price",
  "max_price", "min_price", "unique_departments",
  "avg_price_per_item", "items_per_department", "max_price_share",
  "expected_total_price_cashier", "residual_total_price_cashier",
  "avg_department_risk", "max_department_risk", "total_department_risk",
  "avg_item_loss_rate", "avg_revenue_loss_rate",
  "high_risk_item_share", "high_risk_price_share", "risk_weighted_price",
  dept_feature_names
)

selfcheckout_feature_names <- c(
  shared_feature_names,
  "total_time", "avg_time_per_item", "median_time_per_item", "sd_time_per_item",
  "max_time", "min_time",
  "fast_scan_share", "long_pause_share", "repeated_department_share",
  "price_per_second"
)
```

# Method 1 - Cashier-only K-means on shared features

```{r method1-kmeans-cashier}
aligned1 <- align_feature_columns(cashier_features, test_features, shared_feature_names)
scaler1 <- fit_robust_scaler(aligned1$reference)

cashier_scaled1 <- apply_robust_scaler(aligned1$reference, scaler1)
test_scaled1    <- apply_robust_scaler(aligned1$new, scaler1)

kmeans_cashier <- kmeans(cashier_scaled1, centers = k_kmeans_cashier, nstart = 30, iter.max = 100)
dist1 <- row_distance_to_centers(test_scaled1, kmeans_cashier$centers)

method1_results <- test_features %>%
  transmute(
    transaction_id,
    score_kmeans_cashier = apply(dist1, 1, min),
    cluster_kmeans_cashier = apply(dist1, 1, which.min)
  )
```

# Method 2 - Size-segmented cashier clustering

This method compares each test basket only with cashier baskets of similar basket size.

```{r method2-segmented}
segment_results <- list()

for (bin_name in unique(cashier_features$size_bin)) {
  cashier_bin <- cashier_features %>% filter(size_bin == bin_name)
  test_bin    <- test_features %>% filter(size_bin == bin_name)

  if (nrow(cashier_bin) == 0 || nrow(test_bin) == 0) next

  feature_names_bin <- shared_feature_names
  aligned_bin <- align_feature_columns(cashier_bin, test_bin, feature_names_bin)
  scaler_bin  <- fit_robust_scaler(aligned_bin$reference)

  cashier_scaled_bin <- apply_robust_scaler(aligned_bin$reference, scaler_bin)
  test_scaled_bin    <- apply_robust_scaler(aligned_bin$new, scaler_bin)

  centers_bin <- min(k_segmented_per_bin, max(2, floor(nrow(cashier_bin) / 200)))
  km_bin <- kmeans(cashier_scaled_bin, centers = centers_bin, nstart = 20, iter.max = 100)

  dist_bin <- row_distance_to_centers(test_scaled_bin, km_bin$centers)

  segment_results[[bin_name]] <- tibble(
    transaction_id = test_bin$transaction_id,
    score_segmented_cashier = apply(dist_bin, 1, min),
    cluster_segmented_cashier = apply(dist_bin, 1, which.min)
  )
}

method2_results <- bind_rows(segment_results)

# fallback in case some bins are missing
method2_results <- test_features %>%
  select(transaction_id) %>%
  left_join(method2_results, by = "transaction_id") %>%
  mutate(
    score_segmented_cashier = replace_na(score_segmented_cashier, max(score_segmented_cashier, na.rm = TRUE)),
    cluster_segmented_cashier = replace_na(cluster_segmented_cashier, -1L)
  )
```

# Method 3 - Combined-source K-means on shared features

```{r method3-combined}
set.seed(params$seed)

supermarket_train_combined <- if (nrow(supermarket_features) > max_rows_supermarket_for_combined) {
  supermarket_features %>% slice_sample(n = max_rows_supermarket_for_combined)
} else {
  supermarket_features
}

cashier_train_combined <- if (nrow(cashier_features) > max_rows_cashier_for_combined) {
  cashier_features %>% slice_sample(n = max_rows_cashier_for_combined)
} else {
  cashier_features
}

combined_train <- bind_rows(cashier_train_combined, supermarket_train_combined)

aligned3 <- align_feature_columns(combined_train, test_features, shared_feature_names)
scaler3 <- fit_robust_scaler(aligned3$reference)

combined_scaled <- apply_robust_scaler(aligned3$reference, scaler3)
test_scaled3    <- apply_robust_scaler(aligned3$new, scaler3)

kmeans_combined <- kmeans(combined_scaled, centers = k_kmeans_combined, nstart = 30, iter.max = 100)
dist3 <- row_distance_to_centers(test_scaled3, kmeans_combined$centers)
assigned3 <- apply(dist3, 1, which.min)

combined_train_clusters <- combined_train %>%
  mutate(cluster = kmeans_combined$cluster) %>%
  count(cluster, source, name = "n") %>%
  group_by(cluster) %>%
  mutate(cluster_total = sum(n), share = n / cluster_total) %>%
  ungroup() %>%
  select(cluster, source, share) %>%
  pivot_wider(names_from = source, values_from = share, values_fill = 0)

cluster_profiles <- combined_train_clusters %>%
  transmute(
    cluster,
    cashier_share = cashier,
    supermarket_share = supermarket
  )

method3_results <- tibble(
  transaction_id = test_features$transaction_id,
  cluster_combined = assigned3,
  distance_combined = apply(dist3, 1, min)
) %>%
  left_join(cluster_profiles, by = c("cluster_combined" = "cluster")) %>%
  mutate(
    cashier_share = replace_na(cashier_share, 0),
    supermarket_share = replace_na(supermarket_share, 0),
    score_combined = safe_rank_score(distance_combined) + 0.75 * safe_rank_score(1 - cashier_share)
  )
```

# Method 4 - Supermarket-only self-checkout clustering with timing features

This method uses only self-checkout baskets, so timing features are meaningful again.

```{r method4-selfcheckout}
set.seed(params$seed)

supermarket_train_self <- if (nrow(supermarket_features) > max_rows_supermarket_for_self) {
  supermarket_features %>% slice_sample(n = max_rows_supermarket_for_self)
} else {
  supermarket_features
}

aligned4 <- align_feature_columns(supermarket_train_self, test_features, selfcheckout_feature_names)
scaler4 <- fit_robust_scaler(aligned4$reference)

supermarket_scaled4 <- apply_robust_scaler(aligned4$reference, scaler4)
test_scaled4        <- apply_robust_scaler(aligned4$new, scaler4)

kmeans_self <- kmeans(supermarket_scaled4, centers = k_kmeans_selfcheckout, nstart = 30, iter.max = 100)
dist4 <- row_distance_to_centers(test_scaled4, kmeans_self$centers)
assigned4 <- apply(dist4, 1, which.min)

cluster_sizes_self <- tibble(
  cluster_selfcheckout = seq_len(k_kmeans_selfcheckout),
  cluster_size_selfcheckout = as.integer(table(factor(kmeans_self$cluster, levels = seq_len(k_kmeans_selfcheckout))))
)

method4_results <- tibble(
  transaction_id = test_features$transaction_id,
  cluster_selfcheckout = assigned4,
  distance_selfcheckout = apply(dist4, 1, min)
) %>%
  left_join(cluster_sizes_self, by = "cluster_selfcheckout") %>%
  mutate(
    cluster_size_selfcheckout = replace_na(cluster_size_selfcheckout, 1L),
    score_selfcheckout = safe_rank_score(distance_selfcheckout) + 0.5 * safe_rank_score(1 / cluster_size_selfcheckout)
  )
```

# Final Ensemble Score

The final score puts more weight on:
- cashier-based shared-feature distance,
- size-segmented cashier distance,
- and self-checkout timing-aware clustering.

```{r ensemble}
all_scores <- method1_results %>%
  left_join(method2_results, by = "transaction_id") %>%
  left_join(method3_results %>% select(transaction_id, score_combined, cluster_combined, cashier_share), by = "transaction_id") %>%
  left_join(method4_results %>% select(transaction_id, score_selfcheckout, cluster_selfcheckout), by = "transaction_id") %>%
  mutate(
    rank_kmeans_cashier = safe_rank_score(score_kmeans_cashier),
    rank_segmented_cashier = safe_rank_score(score_segmented_cashier),
    rank_combined = safe_rank_score(score_combined),
    rank_selfcheckout = safe_rank_score(score_selfcheckout),
    final_ensemble_score =
      0.30 * rank_kmeans_cashier +
      0.30 * rank_segmented_cashier +
      0.15 * rank_combined +
      0.25 * rank_selfcheckout
  ) %>%
  arrange(desc(final_ensemble_score),
          desc(score_segmented_cashier),
          desc(score_kmeans_cashier),
          desc(score_selfcheckout))

head(all_scores %>% select(transaction_id, final_ensemble_score, score_kmeans_cashier, score_segmented_cashier, score_combined, score_selfcheckout), 10)
```

# Select the Predicted Fraud Transaction IDs

```{r final-selection}
n_test <- nrow(test_features)
n_flag <- ceiling(expected_fraud_share * n_test)

predicted_fraud_ids <- all_scores %>%
  slice_head(n = n_flag) %>%
  pull(transaction_id)

cat("Number of rows flagged:", length(predicted_fraud_ids), "\n")
head(predicted_fraud_ids, 20)
```

# Write the Transaction IDs to a Text File

```{r write-output}
writeLines(as.character(predicted_fraud_ids), con = output_rows_file)
cat("Predicted transaction IDs written to:", output_rows_file, "\n")
```

# Final Output Block

```{r print-output, results='asis'}
cat(paste(predicted_fraud_ids, collapse = "\n"))
```

# Diagnostics

```{r diagnostics}
all_scores %>%
  summarise(
    avg_score_kmeans_cashier = mean(score_kmeans_cashier, na.rm = TRUE),
    avg_score_segmented_cashier = mean(score_segmented_cashier, na.rm = TRUE),
    avg_score_combined = mean(score_combined, na.rm = TRUE),
    avg_score_selfcheckout = mean(score_selfcheckout, na.rm = TRUE),
    avg_final_score = mean(final_ensemble_score, na.rm = TRUE)
  )

all_scores %>%
  slice_head(n = min(20, n_flag)) %>%
  left_join(test_features, by = "transaction_id") %>%
  select(
    transaction_id,
    final_ensemble_score,
    total_items,
    total_price,
    total_time,
    avg_time_per_item,
    max_price,
    max_price_share,
    unique_departments,
    residual_total_price_cashier,
    cluster_kmeans_cashier,
    cluster_segmented_cashier,
    cluster_combined,
    cluster_selfcheckout,
    cashier_share
  ) %>%
  arrange(desc(final_ensemble_score))
```

# Notes for Interpretation

- Cashier data is now used only with **shared non-time basket features**.
- Timing variables are used only in the **self-checkout-only clustering method**, where they are meaningful.
- The final method is still unsupervised, so it does **not prove fraud**. It identifies baskets that are least consistent with normal basket structure and self-checkout behavior.
- The first things to tune further are:
  - the cluster counts,
  - the ensemble weights,
  - the basket-size segmentation,
  - and the exact feature sets used by the cashier-based and self-checkout-based methods.
