---
title: "Department & Basket Share Analysis"
author: "Niklas"
date: "`20-04-26`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

analysis looks at shopping behaviour patterns across supermarket transactions. Identifies the most frequently visited departments and calculates what percentage of customer baskets contain items from each department.

## Load Libraries & Data

```{r load}
library(dplyr)
library(ggplot2)

long_data <- read.csv("/Users/niklassandelin/Desktop/long_data.csv", stringsAsFactors = FALSE)

cat("Rows:", nrow(long_data), "\n")
cat("Columns:", paste(names(long_data), collapse = ", "), "\n")
```

## Department Basket Share

A department appears in a basket if at least one item in that transaction belongs to it

```{r analysis}
total_baskets <- n_distinct(long_data$transaction_id)
cat("Total unique baskets:", total_baskets, "\n")

dept_basket_share <- long_data %>%
  group_by(department_name) %>%
  summarise(
    basket_count         = n_distinct(transaction_id),
    item_count           = n(),
    avg_items_per_basket = round(item_count / basket_count, 2)
  ) %>%
  mutate(
    basket_share_pct = round(basket_count / total_baskets * 100, 2)
  ) %>%
  arrange(desc(basket_share_pct))

print(dept_basket_share, n = Inf)
```

## Top 10 Departments

```{r top10}
top10 <- head(dept_basket_share, 10)

for (i in seq_len(nrow(top10))) {
  cat(sprintf("  %2d. %-30s  %.1f%% of baskets\n",
              i,
              top10$department_name[i],
              top10$basket_share_pct[i]))
}
```

## Visualisations

### All Departments Basket Share

```{r plot-all, fig.width=10, fig.height=7}
ggplot(dept_basket_share,
       aes(x = reorder(department_name, basket_share_pct),
           y = basket_share_pct)) +
  geom_col(fill = "#2C7BB6", width = 0.7) +
  geom_text(aes(label = paste0(basket_share_pct, "%")),
            hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(dept_basket_share$basket_share_pct) * 1.12),
                     labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Department Presence Across Customer Baskets",
    subtitle = paste0("n = ", format(total_baskets, big.mark = ","), " unique transactions"),
    x        = NULL,
    y        = "% of Baskets Containing Department"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14),
    panel.grid.major.y = element_blank()
  )
```

### Top 10 Departments

```{r plot-top10, fig.width=9, fig.height=6}
ggplot(top10,
       aes(x = reorder(department_name, basket_share_pct),
           y = basket_share_pct)) +
  geom_col(fill = "#1A7340", width = 0.65) +
  geom_text(aes(label = paste0(basket_share_pct, "%")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, max(top10$basket_share_pct) * 1.12),
                     labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Top 10 Departments by Basket Share",
    subtitle = paste0("n = ", format(total_baskets, big.mark = ","), " unique transactions"),
    x        = NULL,
    y        = "% of Baskets"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 15),
    panel.grid.major.y = element_blank()
  )
```
