---
title: "Fraud Risk Scoring and Baseline"
output: pdf_document
date: "2026-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
source("prepare_data.R")
```

# BASELINE MODEL

```{r}
# 1. Basket-level dataset
basket_data <- long_data_clean %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

# 2. Fast / slow scan thresholds
p5_time  <- quantile(basket_data$avg_time_per_item, 0.05, na.rm = TRUE)
p95_time <- quantile(basket_data$avg_time_per_item, 0.95, na.rm = TRUE)

# 3. Basket size group
basket_data <- basket_data %>%
  mutate(
    fast_flag = avg_time_per_item < p5_time,
    slow_flag = avg_time_per_item > p95_time,

    basket_size_group = case_when(
      total_items <= 10 ~ "small",
      total_items <= 25 ~ "medium",
      TRUE ~ "large"
    )
  )

# 4. Expected time given basket size
model_time <- lm(avg_time_per_item ~ total_items, data = basket_data)

basket_data <- basket_data %>%
  mutate(
    expected_time_given_size = predict(model_time, newdata = basket_data),
    time_deviation = avg_time_per_item - expected_time_given_size
  )

# 5. Expected price given basket size
model_price <- lm(total_price ~ total_items, data = basket_data)

basket_data <- basket_data %>%
  mutate(
    expected_price_given_items = predict(model_price, newdata = basket_data),
    price_deviation = total_price - expected_price_given_items,
    price_residual = residuals(model_price),
    price_concentration = max_price / total_price
  )

# 6. Department co-occurrence score
dept_data <- long_data_clean %>%
  select(transaction_id, department_name) %>%
  distinct()

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  rename(
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

co_occurrence <- pairs %>%
  count(dept_1, dept_2, sort = TRUE) %>%
  mutate(score = n / max(n))

pairs_scored <- pairs %>%
  left_join(co_occurrence, by = c("dept_1", "dept_2"))

co_score <- pairs_scored %>%
  group_by(transaction_id) %>%
  summarise(
    co_occurrence_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

basket_data <- basket_data %>%
  left_join(co_score, by = "transaction_id") %>%
  mutate(
    co_occurrence_score = replace_na(co_occurrence_score, 0)
  )

# 7. Thresholds for risk rules
p5_price  <- quantile(basket_data$total_price, 0.05, na.rm = TRUE)
p95_price <- quantile(basket_data$total_price, 0.95, na.rm = TRUE)

p5_resid  <- quantile(basket_data$price_residual, 0.05, na.rm = TRUE)
p95_resid <- quantile(basket_data$price_residual, 0.95, na.rm = TRUE)

p95_conc <- quantile(basket_data$price_concentration, 0.95, na.rm = TRUE)

p5_co <- quantile(basket_data$co_occurrence_score, 0.05, na.rm = TRUE)

# 8. Rule-based risk score
basket_risk <- basket_data %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,

    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,

    unusually_cheap_for_size = price_residual < p5_resid,
    unusually_expensive_for_size = price_residual > p95_resid,

    high_price_concentration = price_concentration > p95_conc,

    unusual_department_combination = co_occurrence_score < p5_co,

    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination)
  ) %>%
  arrange(desc(risk_score))

# 9. Final baseline score
basket_risk <- basket_risk %>%
  mutate(
    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  ) %>%
  arrange(desc(final_score))

# 10. Top suspicious supermarket baskets according to baseline
top_150_baseline_train <- basket_risk %>%
  slice_head(n = 150)
```

# CASE 3: PURE GRAPH MODEL ON basket_data

```{r}
# 1. Distinct departments per basket

dept_data <- long_data_clean %>%
  select(transaction_id, department_name) %>%
  distinct()

# 2. Build unordered department pairs

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

# 3. Department occurrence counts

dept_counts <- dept_data %>%
  count(department_name, name = "dept_count")

# 4. Pair counts

pair_counts <- pairs %>%
  count(dept_1, dept_2, name = "pair_count")

# 5. Conditional co-occurrence strength
# P(dept_2 | dept_1) = pair_count / count(dept_1)
# P(dept_1 | dept_2) = pair_count / count(dept_2)
# We combine both directions symmetrically.

pair_conditional <- pair_counts %>%
  left_join(
    dept_counts %>% rename(dept_1 = department_name, dept_1_count = dept_count),
    by = "dept_1"
  ) %>%
  left_join(
    dept_counts %>% rename(dept_2 = department_name, dept_2_count = dept_count),
    by = "dept_2"
  ) %>%
  mutate(
    p_2_given_1 = pair_count / dept_1_count,
    p_1_given_2 = pair_count / dept_2_count,
    conditional_strength = (p_2_given_1 + p_1_given_2) / 2,
    conditional_rarity = -log(conditional_strength)
  )

# 6. Basket-level graph score on supermarket
# Higher score = more structurally unusual basket

graph_train <- pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

basket_graph <- basket_data %>%
  left_join(graph_train, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  ) %>%
  arrange(desc(graph_score))

# top 15%
graph_threshold <- quantile(basket_graph$graph_score, 0.85, na.rm = TRUE)

basket_graph <- basket_graph %>%
  mutate(
    graph_flag = graph_score >= graph_threshold
  ) %>%
  arrange(desc(graph_score))

table(basket_graph$graph_flag)

top_150_graph_train <- basket_graph %>%
  slice_head(n = 150)

top_150_graph_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    unique_departments,
    graph_score,
    avg_conditional_strength,
    n_pairs
  ) %>%
  head(20)
```

# APPLY PURE GRAPH MODEL TO CASE3

```{r}
case3_lines <- readLines("~/Desktop/case3.csv", warn = FALSE)

case3_raw <- tibble(
  raw_line = case3_lines
) %>%
  mutate(
    transaction_id = stringr::str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = stringr::str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case3_long <- case3_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

case3_basket <- case3_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

case3_dept_data <- case3_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case3_pairs <- case3_dept_data %>%
  inner_join(case3_dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

case3_graph <- case3_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

case3_basket <- case3_basket %>%
  left_join(case3_graph, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0),
    graph_flag = graph_score >= graph_threshold
  ) %>%
  arrange(desc(graph_score))

top_150_graph_case3 <- case3_basket %>%
  slice_head(n = 150)

write_csv(
  top_150_combined_train,
  "~/Desktop/supermarket case/top_150_combined_train.csv"
)

writeLines(as.character(top_150_graph_case3$transaction_id))

readr::write_csv(
  top_150_graph_case3,
  "~/Desktop/supermarket case/test cases/top_150_graph_case3.csv"
)
```
# CASE 4: BASELINE + GRAPH MODEL

```{r}
# weights: 0.6 baseline / 0.4 graph
# graph based on conditional co-occurrence

library(tidyverse)
library(stringr)
library(readr)

# 1. TRAIN: basket_data already built from supermarket
# assumes these already exist:
# - long_data_clean
# - basket_data
# - model_time
# - model_price
# - p5_time, p95_time
# - p5_price, p95_price
# - p5_resid, p95_resid
# - p95_conc, p5_co

# If not yet present in basket_data:
if (!"price_deviation" %in% names(basket_data)) {
  basket_data <- basket_data %>%
    mutate(
      expected_time_given_size = predict(model_time, newdata = basket_data),
      time_deviation = avg_time_per_item - expected_time_given_size,
      expected_price_given_items = predict(model_price, newdata = basket_data),
      price_deviation = total_price - expected_price_given_items,
      price_residual = residuals(model_price),
      price_concentration = max_price / total_price
    )
}

# 2. CONDITIONAL GRAPH MODEL ON SUPERMARKET

dept_data <- long_data_clean %>%
  select(transaction_id, department_name) %>%
  distinct()

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

dept_counts <- dept_data %>%
  count(department_name, name = "dept_count")

pair_counts <- pairs %>%
  count(dept_1, dept_2, name = "pair_count")

pair_conditional <- pair_counts %>%
  left_join(
    dept_counts %>% rename(dept_1 = department_name, dept_1_count = dept_count),
    by = "dept_1"
  ) %>%
  left_join(
    dept_counts %>% rename(dept_2 = department_name, dept_2_count = dept_count),
    by = "dept_2"
  ) %>%
  mutate(
    p_2_given_1 = pair_count / dept_1_count,
    p_1_given_2 = pair_count / dept_2_count,
    conditional_strength = (p_2_given_1 + p_1_given_2) / 2,
    conditional_rarity = -log(conditional_strength)
  )

graph_train <- pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

basket_graph <- basket_data %>%
  left_join(graph_train, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  )

graph_threshold <- quantile(basket_graph$graph_score, 0.85, na.rm = TRUE)

# 3. BASELINE SCORE ON SUPERMARKET
# old baseline version used before combined IF

basket_risk <- basket_graph %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,

    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,

    unusually_cheap_for_size = price_residual < p5_resid,
    unusually_expensive_for_size = price_residual > p95_resid,

    high_price_concentration = price_concentration > p95_conc,
    unusual_department_combination = co_occurrence_score < p5_co,

    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination),

    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  )

# 4. COMBINED SCORE ON SUPERMARKET

basket_risk <- basket_risk %>%
  mutate(
    baseline_score_01 = percent_rank(final_score),
    graph_score_01 = percent_rank(graph_score),
    combined_bg_score = 0.6 * baseline_score_01 + 0.4 * graph_score_01
  ) %>%
  arrange(desc(combined_bg_score))

top_150_baseline_graph_train <- basket_risk %>%
  slice_head(n = 150)

write_csv(
  top_150_baseline_graph_train,
  "~/Desktop/supermarket case/test cases/case4/top_150_baseline_graph_train.csv"
)

top_150_baseline_graph_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_time_per_item,
    price_deviation,
    time_deviation,
    price_concentration,
    co_occurrence_score,
    graph_score,
    risk_score,
    final_score,
    combined_bg_score
  ) %>%
  head(20)
```

# APPLY BASELINE + GRAPH TO CASE4

```{r}
case4_lines <- readLines("~/Desktop/supermarket case/test cases/case4/case4.csv", warn = FALSE)

case4_raw <- tibble(
  raw_line = case4_lines
) %>%
  mutate(
    transaction_id = str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case4_long <- case4_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

case4_basket <- case4_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  )

case4_dept_data <- case4_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case4_pairs <- case4_dept_data %>%
  inner_join(case4_dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

case4_co_score <- case4_pairs %>%
  left_join(
    co_occurrence %>% select(dept_1, dept_2, score),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    co_occurrence_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

case4_graph <- case4_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

case4_basket <- case4_basket %>%
  left_join(case4_co_score, by = "transaction_id") %>%
  left_join(case4_graph, by = "transaction_id") %>%
  mutate(
    co_occurrence_score = replace_na(co_occurrence_score, 0),
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0),

    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,

    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items,

    price_concentration = max_price / total_price
  )

case4_basket <- case4_basket %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,

    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,

    unusually_cheap_for_size = price_deviation < p5_resid,
    unusually_expensive_for_size = price_deviation > p95_resid,

    high_price_concentration = price_concentration > p95_conc,
    unusual_department_combination = co_occurrence_score < p5_co,

    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination),

    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  ) 

case4_basket <- case4_basket %>%
  mutate(
    baseline_score_01 = percent_rank(final_score),
    graph_score_01 = percent_rank(graph_score),
    combined_bg_score = 0.6 * baseline_score_01 + 0.4 * graph_score_01
  ) %>%
  arrange(desc(combined_bg_score))

top_150_baseline_graph_case4 <- case4_basket %>%
  slice_head(n = 150)

# submission list
writeLines(as.character(top_150_baseline_graph_case4$transaction_id))

# optional save
readr::write_csv(
  top_150_baseline_graph_case4,
  "~/Desktop/supermarket case/test cases/case4/top_150_baseline_graph_case4.csv"
)
```
# CASE 5: IF + GRAPH

```{r}
# IF + GRAPH ON SUPERMARKET
# weights: 0.6 IF / 0.4 graph

library(tidyverse)
library(isotree)
library(readr)
library(stringr)

basket_data <- long_data_clean %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    max_time = max(time, na.rm = TRUE),
    min_time = min(time, na.rm = TRUE),
    price_per_second = ifelse(sum(time, na.rm = TRUE) > 0, sum(price, na.rm = TRUE) / sum(time, na.rm = TRUE), NA_real_),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  ) %>%
  mutate(
    max_price_share = ifelse(total_price > 0, max_price / total_price, NA_real_)
  )


# If needed:
if (!"max_price_share" %in% names(basket_data)) {
  basket_data <- basket_data %>%
    mutate(
      max_price_share = ifelse(total_price > 0, max_price / total_price, NA_real_)
    )
}

# B. PURE IF ON basket_data

if_train <- basket_data %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    max_price,
    max_price_share,
    unique_departments
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(price_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(max_price_share, na.rm = TRUE))
  )

X_train_raw <- if_train %>%
  select(-transaction_id)

X_train <- scale(X_train_raw)

if_model <- isolation.forest(
  as.matrix(X_train),
  ntrees = 300,
  sample_size = 512,
  ndim = 1,
  nthreads = 1
)

if_train$if_score <- predict(
  if_model,
  as.matrix(X_train),
  type = "score"
)

# C. PURE GRAPH ON supermarket
# conditional co-occurrence

dept_data <- long_data_clean %>%
  select(transaction_id, department_name) %>%
  distinct()

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

dept_counts <- dept_data %>%
  count(department_name, name = "dept_count")

pair_counts <- pairs %>%
  count(dept_1, dept_2, name = "pair_count")

pair_conditional <- pair_counts %>%
  left_join(
    dept_counts %>% rename(dept_1 = department_name, dept_1_count = dept_count),
    by = "dept_1"
  ) %>%
  left_join(
    dept_counts %>% rename(dept_2 = department_name, dept_2_count = dept_count),
    by = "dept_2"
  ) %>%
  mutate(
    p_2_given_1 = pair_count / dept_1_count,
    p_1_given_2 = pair_count / dept_2_count,
    conditional_strength = (p_2_given_1 + p_1_given_2) / 2,
    conditional_rarity = -log(conditional_strength)
  )

graph_train <- pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

train_ig <- basket_data %>%
  left_join(if_train %>% select(transaction_id, if_score), by = "transaction_id") %>%
  left_join(graph_train, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  ) %>%
  mutate(
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    combined_ig_score = 0.6 * if_score_01 + 0.4 * graph_score_01
  ) %>%
  arrange(desc(combined_ig_score))

top_150_if_graph_train <- train_ig %>%
  slice_head(n = 150)


top_150_if_graph_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_time_per_item,
    max_price_share,
    unique_departments,
    if_score,
    graph_score,
    combined_ig_score
  ) %>%
  head(20)
```

# APPLY IF + GRAPH TO CASE 5 (NULL GRAPH SCORE FOR ALL)

```{r}
case5_lines <- readLines("~/Desktop/case5.csv", warn = FALSE)

case5_raw <- tibble(
  raw_line = case5_lines
) %>%
  mutate(
    transaction_id = str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case5_long <- case5_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

case5_basket <- case5_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(sum(time, na.rm = TRUE) > 0, sum(price, na.rm = TRUE) / sum(time, na.rm = TRUE), NA_real_),
    max_price = max(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  ) %>%
  mutate(
    max_price_share = ifelse(total_price > 0, max_price / total_price, NA_real_)
  )

# IF features
if_case5 <- case5_basket %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    max_price,
    max_price_share,
    unique_departments
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(if_train$sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(if_train$price_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(if_train$max_price_share, na.rm = TRUE))
  )

X_case5_raw <- if_case5 %>%
  select(-transaction_id)

X_case5 <- scale(
  X_case5_raw,
  center = attr(X_train, "scaled:center"),
  scale = attr(X_train, "scaled:scale")
)

if_case5$if_score <- predict(
  if_model,
  as.matrix(X_case5),
  type = "score"
)

# Graph features
case5_dept_data <- case5_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case5_pairs <- case5_dept_data %>%
  inner_join(case3_dept_data, by = "transaction_id", relationship = "many-to-many") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

case5_graph <- case5_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

case5_ig <- case5_basket %>%
  left_join(if_case5 %>% select(transaction_id, if_score), by = "transaction_id") %>%
  left_join(case5_graph, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  ) %>%
  mutate(
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    combined_ig_score = 0.6 * if_score_01 + 0.4 * graph_score_01
  ) %>%
  arrange(desc(combined_ig_score))

top_150_if_graph_case5 <- case5_ig %>%
  slice_head(n = 150)

writeLines(as.character(top_150_if_graph_case5$transaction_id))

# optional save
readr::write_csv(
  top_150_if_graph_case5,
  "~/Desktop/supermarket case/test cases/top_150_if_graph_case5.csv"
)
```

#PREPARATION CASE 6: IF + BASELINE + GRAPH

```{r}
# 1. READ + CLEAN TRAIN DATA

lines <- readLines("~/Desktop/supermarket case/supermarket.csv", warn = FALSE)

supermarket_raw <- tibble(
  transaction_id = 1:length(lines),
  basket = lines
)

dept_map <- c(
  "1" = "Bakery & Pastry",
  "2" = "Beer & Wine",
  "3" = "Books & Magazines",
  "4" = "Candy & Chips",
  "5" = "Care & Hygiene",
  "6" = "Cereals & Spreads",
  "7" = "Cheese & Tapas",
  "8" = "Dairy & Eggs",
  "9" = "Freezer",
  "10" = "Fruit & Vegetables",
  "11" = "Household & Pet",
  "12" = "Meat & Fish",
  "13" = "Pasta & Rice",
  "14" = "Salads & Meals",
  "15" = "Sauces & Spices",
  "16" = "Soda & Juices",
  "17" = "Special Diet",
  "18" = "Vegetarian & Vegan"
)

long_data <- supermarket_raw %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  )

long_data_clean <- long_data %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  ) %>%
  mutate(
    price_per_second_item = ifelse(time > 0, price / time, NA_real_)
  )

# 2. FULL BASKET DATASET

basket_data_full <- long_data_clean %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    max_time = max(time, na.rm = TRUE),
    min_time = min(time, na.rm = TRUE),
    price_per_second = ifelse(sum(time, na.rm = TRUE) > 0,
                              sum(price, na.rm = TRUE) / sum(time, na.rm = TRUE),
                              NA_real_),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  ) %>%
  mutate(
    max_price_share = ifelse(total_price > 0, max_price / total_price, NA_real_),
    items_per_second = ifelse(total_time > 0, total_items / total_time, NA_real_),
    department_diversity_ratio = unique_departments / total_items
  )

dept_share <- long_data_clean %>%
  count(transaction_id, department, name = "dept_items") %>%
  group_by(transaction_id) %>%
  summarise(
    dominant_department_share = max(dept_items) / sum(dept_items),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(dept_share, by = "transaction_id")

# 3. TIME + PRICE DEVIATION MODELS

model_time <- lm(avg_time_per_item ~ total_items, data = basket_data_full)
model_price <- lm(total_price ~ total_items, data = basket_data_full)

basket_data_full <- basket_data_full %>%
  mutate(
    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,
    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items,
    price_residual = residuals(model_price),
    price_concentration = ifelse(total_price > 0, max_price / total_price, NA_real_)
  )

# 4. OLD CO-OCCURRENCE SCORE
# needed for baseline

dept_data <- long_data_clean %>%
  select(transaction_id, department_name) %>%
  distinct()

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

co_occurrence <- pairs %>%
  count(dept_1, dept_2, sort = TRUE) %>%
  mutate(score = n / max(n))

co_score <- pairs %>%
  left_join(co_occurrence %>% select(dept_1, dept_2, score),
            by = c("dept_1", "dept_2")) %>%
  group_by(transaction_id) %>%
  summarise(
    co_occurrence_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(co_score, by = "transaction_id") %>%
  mutate(
    co_occurrence_score = replace_na(co_occurrence_score, 0)
  )

# 5. CONDITIONAL GRAPH SCORE
# needed for graph model

dept_counts <- dept_data %>%
  count(department_name, name = "dept_count")

pair_counts <- pairs %>%
  count(dept_1, dept_2, name = "pair_count")

pair_conditional <- pair_counts %>%
  left_join(
    dept_counts %>% rename(dept_1 = department_name, dept_1_count = dept_count),
    by = "dept_1"
  ) %>%
  left_join(
    dept_counts %>% rename(dept_2 = department_name, dept_2_count = dept_count),
    by = "dept_2"
  ) %>%
  mutate(
    p_2_given_1 = pair_count / dept_1_count,
    p_1_given_2 = pair_count / dept_2_count,
    conditional_strength = (p_2_given_1 + p_1_given_2) / 2,
    conditional_rarity = -log(conditional_strength)
  )

graph_train <- pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(graph_train, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  )

# 6. BASELINE THRESHOLDS

p5_time  <- quantile(basket_data_full$avg_time_per_item, 0.05, na.rm = TRUE)
p95_time <- quantile(basket_data_full$avg_time_per_item, 0.95, na.rm = TRUE)

p5_price  <- quantile(basket_data_full$total_price, 0.05, na.rm = TRUE)
p95_price <- quantile(basket_data_full$total_price, 0.95, na.rm = TRUE)

p5_resid  <- quantile(basket_data_full$price_residual, 0.05, na.rm = TRUE)
p95_resid <- quantile(basket_data_full$price_residual, 0.95, na.rm = TRUE)

p95_conc <- quantile(basket_data_full$price_concentration, 0.95, na.rm = TRUE)
p5_co    <- quantile(basket_data_full$co_occurrence_score, 0.05, na.rm = TRUE)

graph_threshold <- quantile(basket_data_full$graph_score, 0.85, na.rm = TRUE)

# 7. SAFE COPIES FOR DIFFERENT MODELS

basket_data_baseline <- basket_data_full
basket_data_if <- basket_data_full
basket_data_graph <- basket_data_full
```

# CASE 6:

```{r}
library(tidyverse)
library(isotree)

# 1. BASELINE SCORE

baseline_train <- basket_data_full %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,

    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,

    unusually_cheap_for_size = price_residual < p5_resid,
    unusually_expensive_for_size = price_residual > p95_resid,

    high_price_concentration = price_concentration > p95_conc,
    unusual_department_combination = co_occurrence_score < p5_co,

    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination),

    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  )

# 2. IF SCORE

if_train <- baseline_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    items_per_second,
    max_price,
    max_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    price_deviation,
    time_deviation
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(max_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(dominant_department_share, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(time_deviation, na.rm = TRUE))
  )

X_train_raw <- if_train %>%
  select(-transaction_id)

X_train <- scale(X_train_raw)

if_model <- isolation.forest(
  as.matrix(X_train),
  ntrees = 300,
  sample_size = 512,
  ndim = 1,
  nthreads = 1
)

if_train$if_score <- predict(
  if_model,
  as.matrix(X_train),
  type = "score"
)

# 3. MERGE BASELINE + IF + GRAPH

combined_train <- baseline_train %>%
  left_join(
    if_train %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  mutate(
    baseline_score_01 = percent_rank(final_score),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),

    combined_big_score =
      0.5 * baseline_score_01 +
      0.3 * if_score_01 +
      0.2 * graph_score_01
  ) %>%
  arrange(desc(combined_big_score))

# 4. TOP 150 ON TRAIN

top_150_combined_big_train <- combined_train %>%
  slice_head(n = 150)

top_150_combined_big_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_time_per_item,
    price_deviation,
    time_deviation,
    co_occurrence_score,
    graph_score,
    risk_score,
    final_score,
    if_score,
    combined_big_score
  ) %>%
  head(20)

readr::write_csv(
  top_150_case6,
  "~/Desktop/supermarket case/test cases/case6/top_150_baseline_if_graph_train.csv"
)
```

# APPLY FOR CASE 6

```{r}
# CASE5: BASELINE + IF + GRAPH
# weights: 0.5 baseline / 0.3 IF / 0.2 graph

library(tidyverse)
library(stringr)
library(readr)

case6_lines <- readLines("~/Desktop/supermarket case/test cases/case6/case6.csv", warn = FALSE)

case6_raw <- tibble(
  raw_line = case6_lines
) %>%
  mutate(
    transaction_id = str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case6_long <- case6_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

case6_basket <- case6_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    max_time = max(time, na.rm = TRUE),
    min_time = min(time, na.rm = TRUE),
    price_per_second = ifelse(total_time > 0, total_price / total_time, NA_real_),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    .groups = "drop"
  ) %>%
  mutate(
    max_price_share = ifelse(total_price > 0, max_price / total_price, NA_real_),
    items_per_second = ifelse(total_time > 0, total_items / total_time, NA_real_),
    department_diversity_ratio = unique_departments / total_items
  )

case6_dept_share <- case6_long %>%
  count(transaction_id, department, name = "dept_items") %>%
  group_by(transaction_id) %>%
  summarise(
    dominant_department_share = max(dept_items) / sum(dept_items),
    .groups = "drop"
  )

case6_dept_data <- case6_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case6_pairs <- case6_dept_data %>%
  inner_join(case6_dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

# old co-occurrence score for baseline
case6_co_score <- case6_pairs %>%
  left_join(
    co_occurrence %>% select(dept_1, dept_2, score),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    co_occurrence_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

# conditional graph score
case6_graph <- case6_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

case6_basket <- case6_basket %>%
  left_join(case6_dept_share, by = "transaction_id") %>%
  left_join(case6_co_score, by = "transaction_id") %>%
  left_join(case6_graph, by = "transaction_id") %>%
  mutate(
    dominant_department_share = replace_na(dominant_department_share, 1),
    co_occurrence_score = replace_na(co_occurrence_score, 0),
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0),

    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,

    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items,

    price_concentration = ifelse(total_price > 0, max_price / total_price, NA_real_)
  )

# baseline
case6_basket <- case6_basket %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,

    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,

    unusually_cheap_for_size = price_deviation < p5_resid,
    unusually_expensive_for_size = price_deviation > p95_resid,

    high_price_concentration = price_concentration > p95_conc,
    unusual_department_combination = co_occurrence_score < p5_co,

    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination),

    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  )

# IF
if_case6 <- case6_basket %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    items_per_second,
    max_price,
    max_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    price_deviation,
    time_deviation
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(if_train$sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(if_train$price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(if_train$items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(if_train$max_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(if_train$department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(if_train$dominant_department_share, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(if_train$price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(if_train$time_deviation, na.rm = TRUE))
  )

X_case6_raw <- if_case6 %>%
  select(-transaction_id)

X_case6 <- scale(
  X_case6_raw,
  center = attr(X_train, "scaled:center"),
  scale = attr(X_train, "scaled:scale")
)

if_case6$if_score <- predict(
  if_model,
  as.matrix(X_case6),
  type = "score"
)

# combine all 3
case6_basket <- case6_basket %>%
  left_join(
    if_case6 %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  mutate(
    baseline_score_01 = percent_rank(final_score),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    combined_big_score =
      0.5 * baseline_score_01 +
      0.3 * if_score_01 +
      0.2 * graph_score_01
  ) %>%
  arrange(desc(combined_big_score))

top_150_case6 <- case6_basket %>%
  slice_head(n = 150)

writeLines(as.character(top_150_case6$transaction_id))

# optional save
readr::write_csv(
  top_150_case6,
  "~/Desktop/supermarket case/test cases/case6/top_150_case6_baseline_if_graph.csv"
)
```


# 1. CASE 5 DIAGNOSTICS 

```{r}

case5_lines <- readLines("~/Desktop/case5.csv", warn = FALSE)

case5_raw <- tibble(
  raw_line = case5_lines
) %>%
  mutate(
    transaction_id = str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case5_long <- case5_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

# 2. distinct departments

case5_dept_data <- case5_long %>%
  select(transaction_id, department_name) %>%
  distinct()

cat("rows in case5_dept_data:", nrow(case5_dept_data), "\n")
cat("missing department names:", sum(is.na(case5_dept_data$department_name)), "\n\n")

print(
  case5_dept_data %>%
    count(department_name, sort = TRUE)
)

# 3. build graph pairs

case5_pairs <- case5_dept_data %>%
  inner_join(case5_dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

cat("\nrows in case5_pairs:", nrow(case5_pairs), "\n")
cat("unique transactions with pairs:", n_distinct(case5_pairs$transaction_id), "\n\n")

print(head(case5_pairs, 20))

# 4. join with train graph

case5_graph_debug <- case5_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  )

cat("\nrows after join:", nrow(case5_graph_debug), "\n")
cat("missing conditional_strength:", sum(is.na(case5_graph_debug$conditional_strength)), "\n")
cat("missing conditional_rarity:", sum(is.na(case5_graph_debug$conditional_rarity)), "\n")
cat("share missing rarity:", mean(is.na(case5_graph_debug$conditional_rarity)), "\n\n")

print(head(case5_graph_debug, 20))

# 5. unmatched pairs

unmatched_pairs <- case5_graph_debug %>%
  filter(is.na(conditional_rarity)) %>%
  count(dept_1, dept_2, sort = TRUE)

cat("\nnumber of unmatched pair types:", nrow(unmatched_pairs), "\n\n")
print(head(unmatched_pairs, 30))

# 6. matched pairs

matched_pairs <- case5_graph_debug %>%
  filter(!is.na(conditional_rarity))

cat("\nmatched rows:", nrow(matched_pairs), "\n")

if (nrow(matched_pairs) > 0) {
  print(summary(matched_pairs$conditional_rarity))
}

# 7. basket-level graph score

case5_graph <- case5_graph_debug %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    n_matched_pairs = sum(!is.na(conditional_rarity)),
    .groups = "drop"
  )

cat("\nsummary of basket-level graph_score:\n")
print(summary(case5_graph$graph_score))

cat("\ntransactions with zero matched pairs:\n")
print(
  case5_graph %>%
    filter(n_matched_pairs == 0) %>%
    select(transaction_id, n_pairs, n_matched_pairs) %>%
    head(30)
)
```


# CASE 5 DEBAG -> CASE 7

```{r}

train_lines <- readLines("~/Desktop/supermarket case/supermarket.csv", warn = FALSE)

train_raw <- tibble(
  transaction_id = 1:length(train_lines),
  basket = train_lines
)

dept_map <- c(
  "1" = "Bakery & Pastry",
  "2" = "Beer & Wine",
  "3" = "Books & Magazines",
  "4" = "Candy & Chips",
  "5" = "Care & Hygiene",
  "6" = "Cereals & Spreads",
  "7" = "Cheese & Tapas",
  "8" = "Dairy & Eggs",
  "9" = "Freezer",
  "10" = "Fruit & Vegetables",
  "11" = "Household & Pet",
  "12" = "Meat & Fish",
  "13" = "Pasta & Rice",
  "14" = "Salads & Meals",
  "15" = "Sauces & Spices",
  "16" = "Soda & Juices",
  "17" = "Special Diet",
  "18" = "Vegetarian & Vegan"
)

train_long <- train_raw %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

basket_data_full <- train_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(total_time > 0, total_price / total_time, NA_real_),
    max_price = max(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    max_price_share = ifelse(total_price > 0, max(price, na.rm = TRUE) / total_price, NA_real_),
    .groups = "drop"
  )

dept_share <- train_long %>%
  count(transaction_id, department, name = "dept_items") %>%
  group_by(transaction_id) %>%
  summarise(
    dominant_department_share = max(dept_items) / sum(dept_items),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(dept_share, by = "transaction_id") %>%
  mutate(
    items_per_second = ifelse(total_time > 0, total_items / total_time, NA_real_),
    department_diversity_ratio = unique_departments / total_items
  )

model_time <- lm(avg_time_per_item ~ total_items, data = basket_data_full)
model_price <- lm(total_price ~ total_items, data = basket_data_full)

basket_data_full <- basket_data_full %>%
  mutate(
    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,
    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items
  )

if_train <- basket_data_full %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    items_per_second,
    max_price,
    max_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    price_deviation,
    time_deviation
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(max_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(dominant_department_share, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(time_deviation, na.rm = TRUE))
  )

X_train_raw <- if_train %>%
  select(-transaction_id)

X_train <- scale(X_train_raw)

if_model <- isolation.forest(
  as.matrix(X_train),
  ntrees = 300,
  sample_size = 512,
  ndim = 1,
  nthreads = 1
)

# graph train
dept_data <- train_long %>%
  select(transaction_id, department_name) %>%
  distinct()

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id", relationship = "many-to-many") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

dept_counts <- dept_data %>%
  count(department_name, name = "dept_count")

pair_counts <- pairs %>%
  count(dept_1, dept_2, name = "pair_count")

pair_conditional <- pair_counts %>%
  left_join(
    dept_counts %>% rename(dept_1 = department_name, dept_1_count = dept_count),
    by = "dept_1"
  ) %>%
  left_join(
    dept_counts %>% rename(dept_2 = department_name, dept_2_count = dept_count),
    by = "dept_2"
  ) %>%
  mutate(
    p_2_given_1 = pair_count / dept_1_count,
    p_1_given_2 = pair_count / dept_2_count,
    conditional_strength = (p_2_given_1 + p_1_given_2) / 2,
    conditional_rarity = -log(conditional_strength)
  )

summary(top_150_if_graph_case7$graph_score)
summary(top_150_if_graph_case7$graph_score_01)
summary(top_150_if_graph_case7$if_score)

writeLines(as.character(top_150_if_graph_case7$transaction_id))

readr::write_csv(
  top_150_if_graph_train,
  "~/Desktop/supermarket case/test cases/top_150_if_graph_train.csv"
)

```


```{r}
case7_lines <- readLines("~/Desktop/case7.csv", warn = FALSE)

case7_raw <- tibble(
  raw_line = case7_lines
) %>%
  mutate(
    transaction_id = stringr::str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = stringr::str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case7_long <- case7_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

case7_basket <- case7_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(total_time > 0, total_price / total_time, NA_real_),
    max_price = max(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    max_price_share = ifelse(total_price > 0, max(price, na.rm = TRUE) / total_price, NA_real_),
    .groups = "drop"
  )

case7_dept_share <- case7_long %>%
  count(transaction_id, department, name = "dept_items") %>%
  group_by(transaction_id) %>%
  summarise(
    dominant_department_share = max(dept_items) / sum(dept_items),
    .groups = "drop"
  )

case7_basket <- case7_basket %>%
  left_join(case7_dept_share, by = "transaction_id") %>%
  mutate(
    items_per_second = ifelse(total_time > 0, total_items / total_time, NA_real_),
    department_diversity_ratio = unique_departments / total_items
  )

# deviations for IF

case7_basket <- case7_basket %>%
  mutate(
    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,
    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items
  )

# IF features

if_case7 <- case7_basket %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    items_per_second,
    max_price,
    max_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    price_deviation,
    time_deviation
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(if_train$sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(if_train$price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(if_train$items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(if_train$max_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(if_train$department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(if_train$dominant_department_share, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(if_train$price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(if_train$time_deviation, na.rm = TRUE))
  )

X_case7_raw <- if_case7 %>%
  select(-transaction_id)

X_case7 <- scale(
  X_case7_raw,
  center = attr(X_train, "scaled:center"),
  scale = attr(X_train, "scaled:scale")
)

if_case7$if_score <- predict(
  if_model,
  as.matrix(X_case7),
  type = "score"
)

# graph features

case7_dept_data <- case7_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case7_pairs <- case7_dept_data %>%
  inner_join(case7_dept_data, by = "transaction_id", relationship = "many-to-many") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

case7_graph <- case7_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

# combine IF + graph

case7_ig <- case7_basket %>%
  left_join(
    if_case7 %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  left_join(
    case7_graph,
    by = "transaction_id"
  ) %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    combined_ig_score = 0.6 * if_score_01 + 0.4 * graph_score_01
  ) %>%
  arrange(desc(combined_ig_score))

top_150_if_graph_case7 <- case7_ig %>%
  slice_head(n = 150)

summary(top_150_if_graph_case7$graph_score)
summary(top_150_if_graph_case7$graph_score_01)
summary(top_150_if_graph_case7$if_score)

writeLines(as.character(top_150_if_graph_case7$transaction_id))

readr::write_csv(
  top_150_if_graph_case7,
  "~/Desktop/supermarket case/test cases/top_150_if_graph_case7.csv"
)
```

# CASE 8 -> RE-DO CASE 6

```{r}
# TRAIN DATA

train_lines <- readLines("~/Desktop/supermarket case/supermarket.csv", warn = FALSE)

train_raw <- tibble(
  transaction_id = 1:length(train_lines),
  basket = train_lines
)

dept_map <- c(
  "1" = "Bakery & Pastry",
  "2" = "Beer & Wine",
  "3" = "Books & Magazines",
  "4" = "Candy & Chips",
  "5" = "Care & Hygiene",
  "6" = "Cereals & Spreads",
  "7" = "Cheese & Tapas",
  "8" = "Dairy & Eggs",
  "9" = "Freezer",
  "10" = "Fruit & Vegetables",
  "11" = "Household & Pet",
  "12" = "Meat & Fish",
  "13" = "Pasta & Rice",
  "14" = "Salads & Meals",
  "15" = "Sauces & Spices",
  "16" = "Soda & Juices",
  "17" = "Special Diet",
  "18" = "Vegetarian & Vegan"
)

train_long <- train_raw %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

# FULL BASKET TABLE

basket_data_full <- train_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(total_time > 0, total_price / total_time, NA_real_),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    max_price_share = ifelse(total_price > 0, max(price, na.rm = TRUE) / total_price, NA_real_),
    .groups = "drop"
  )

dept_share <- train_long %>%
  count(transaction_id, department, name = "dept_items") %>%
  group_by(transaction_id) %>%
  summarise(
    dominant_department_share = max(dept_items) / sum(dept_items),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(dept_share, by = "transaction_id") %>%
  mutate(
    items_per_second = ifelse(total_time > 0, total_items / total_time, NA_real_),
    department_diversity_ratio = unique_departments / total_items
  )

# BASELINE MODELS

model_time <- lm(avg_time_per_item ~ total_items, data = basket_data_full)
model_price <- lm(total_price ~ total_items, data = basket_data_full)

basket_data_full <- basket_data_full %>%
  mutate(
    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,
    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items,
    price_residual = residuals(model_price),
    price_concentration = ifelse(total_price > 0, max_price / total_price, NA_real_)
  )

# OLD CO-OCCURRENCE SCORE FOR BASELINE

dept_data <- train_long %>%
  select(transaction_id, department_name) %>%
  distinct()

pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id", relationship = "many-to-many") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

co_occurrence <- pairs %>%
  count(dept_1, dept_2, sort = TRUE) %>%
  mutate(score = n / max(n))

co_score <- pairs %>%
  left_join(
    co_occurrence %>% select(dept_1, dept_2, score),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    co_occurrence_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(co_score, by = "transaction_id") %>%
  mutate(
    co_occurrence_score = replace_na(co_occurrence_score, 0)
  )

# CONDITIONAL GRAPH SCORE

dept_counts <- dept_data %>%
  count(department_name, name = "dept_count")

pair_counts <- pairs %>%
  count(dept_1, dept_2, name = "pair_count")

pair_conditional <- pair_counts %>%
  left_join(
    dept_counts %>% rename(dept_1 = department_name, dept_1_count = dept_count),
    by = "dept_1"
  ) %>%
  left_join(
    dept_counts %>% rename(dept_2 = department_name, dept_2_count = dept_count),
    by = "dept_2"
  ) %>%
  mutate(
    p_2_given_1 = pair_count / dept_1_count,
    p_1_given_2 = pair_count / dept_2_count,
    conditional_strength = (p_2_given_1 + p_1_given_2) / 2,
    conditional_rarity = -log(conditional_strength)
  )

graph_train <- pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

basket_data_full <- basket_data_full %>%
  left_join(graph_train, by = "transaction_id") %>%
  mutate(
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  )

# BASELINE THRESHOLDS

p5_time  <- quantile(basket_data_full$avg_time_per_item, 0.05, na.rm = TRUE)
p95_time <- quantile(basket_data_full$avg_time_per_item, 0.95, na.rm = TRUE)

p5_price  <- quantile(basket_data_full$total_price, 0.05, na.rm = TRUE)
p95_price <- quantile(basket_data_full$total_price, 0.95, na.rm = TRUE)

p5_resid  <- quantile(basket_data_full$price_residual, 0.05, na.rm = TRUE)
p95_resid <- quantile(basket_data_full$price_residual, 0.95, na.rm = TRUE)

p95_conc <- quantile(basket_data_full$price_concentration, 0.95, na.rm = TRUE)
p5_co    <- quantile(basket_data_full$co_occurrence_score, 0.05, na.rm = TRUE)

# BASELINE SCORE ON TRAIN

baseline_train <- basket_data_full %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,
    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,
    unusually_cheap_for_size = price_residual < p5_resid,
    unusually_expensive_for_size = price_residual > p95_resid,
    high_price_concentration = price_concentration > p95_conc,
    unusual_department_combination = co_occurrence_score < p5_co,
    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination),
    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  )

# IF TRAIN

if_train <- baseline_train %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    items_per_second,
    max_price,
    max_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    price_deviation,
    time_deviation
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(max_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(dominant_department_share, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(time_deviation, na.rm = TRUE))
  )

X_train_raw <- if_train %>%
  select(-transaction_id)

X_train <- scale(X_train_raw)

if_model <- isolation.forest(
  as.matrix(X_train),
  ntrees = 300,
  sample_size = 512,
  ndim = 1,
  nthreads = 1
)

if_train$if_score <- predict(
  if_model,
  as.matrix(X_train),
  type = "score"
)

baseline_train <- baseline_train %>%
  left_join(
    if_train %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  mutate(
    baseline_score_01 = percent_rank(final_score),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    combined_big_score =
      0.25 * baseline_score_01 +
      0.45 * if_score_01 +
      0.30 * graph_score_01
  ) %>%
  arrange(desc(combined_big_score))

top_150_train_big <- baseline_train %>%
  slice_head(n = 150)

summary(top_150_train_big$final_score)
summary(top_150_train_big$if_score)
summary(top_150_train_big$graph_score)

write.csv(
  top_150_train_big,
  "~/Desktop/supermarket case/test cases/top_150_train_baseline_if_graph.csv",
  row.names = FALSE
)
```
# APPLY ON CASE 8:

```{r}
# CASE8: BASELINE + IF + GRAPH
# weights:
# baseline 0.25
# IF       0.45
# graph    0.30

case8_lines <- readLines("~/Desktop/case8.csv", warn = FALSE)

case8_raw <- tibble(
  raw_line = case8_lines
) %>%
  mutate(
    transaction_id = str_extract(raw_line, "(?<=Transaction\\s)\\d+"),
    transaction_id = as.integer(transaction_id),
    basket = str_remove(raw_line, "^Transaction\\s+\\d+:")
  )

case8_long <- case8_raw %>%
  select(transaction_id, basket) %>%
  separate_rows(basket, sep = ",") %>%
  mutate(basket = str_trim(basket)) %>%
  separate(
    basket,
    into = c("department", "time", "price"),
    sep = "\\s+",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(
    department = as.integer(department),
    time = as.numeric(time),
    price = as.numeric(price),
    department_name = dept_map[as.character(department)]
  ) %>%
  filter(
    !is.na(department),
    !is.na(time),
    !is.na(price),
    time >= 0,
    price >= 0
  )

case8_basket <- case8_long %>%
  group_by(transaction_id) %>%
  summarise(
    total_items = n(),
    total_price = sum(price, na.rm = TRUE),
    avg_price = mean(price, na.rm = TRUE),
    total_time = sum(time, na.rm = TRUE),
    avg_time_per_item = mean(time, na.rm = TRUE),
    sd_time_per_item = sd(time, na.rm = TRUE),
    price_per_second = ifelse(total_time > 0, total_price / total_time, NA_real_),
    max_price = max(price, na.rm = TRUE),
    min_price = min(price, na.rm = TRUE),
    unique_departments = n_distinct(department),
    max_price_share = ifelse(total_price > 0, max(price, na.rm = TRUE) / total_price, NA_real_),
    .groups = "drop"
  )

case8_dept_share <- case8_long %>%
  count(transaction_id, department, name = "dept_items") %>%
  group_by(transaction_id) %>%
  summarise(
    dominant_department_share = max(dept_items) / sum(dept_items),
    .groups = "drop"
  )

case8_basket <- case8_basket %>%
  left_join(case8_dept_share, by = "transaction_id") %>%
  mutate(
    items_per_second = ifelse(total_time > 0, total_items / total_time, NA_real_),
    department_diversity_ratio = unique_departments / total_items
  ) %>%
  mutate(
    expected_time_given_size = predict(model_time, newdata = .),
    time_deviation = avg_time_per_item - expected_time_given_size,
    expected_price_given_items = predict(model_price, newdata = .),
    price_deviation = total_price - expected_price_given_items,
    price_concentration = ifelse(total_price > 0, max_price / total_price, NA_real_)
  )

# old co-occurrence score for baseline
case8_dept_data <- case8_long %>%
  select(transaction_id, department_name) %>%
  distinct()

case8_pairs <- case8_dept_data %>%
  inner_join(case8_dept_data, by = "transaction_id", relationship = "many-to-many") %>%
  filter(department_name.x < department_name.y) %>%
  transmute(
    transaction_id,
    dept_1 = department_name.x,
    dept_2 = department_name.y
  )

case8_co_score <- case8_pairs %>%
  left_join(
    co_occurrence %>% select(dept_1, dept_2, score),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    co_occurrence_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

# graph score
case8_graph <- case8_pairs %>%
  left_join(
    pair_conditional %>%
      select(dept_1, dept_2, conditional_strength, conditional_rarity),
    by = c("dept_1", "dept_2")
  ) %>%
  group_by(transaction_id) %>%
  summarise(
    graph_score = mean(conditional_rarity, na.rm = TRUE),
    avg_conditional_strength = mean(conditional_strength, na.rm = TRUE),
    n_pairs = n(),
    .groups = "drop"
  )

case8_basket <- case8_basket %>%
  left_join(case8_co_score, by = "transaction_id") %>%
  left_join(case8_graph, by = "transaction_id") %>%
  mutate(
    dominant_department_share = replace_na(dominant_department_share, 1),
    co_occurrence_score = replace_na(co_occurrence_score, 0),
    graph_score = replace_na(graph_score, 0),
    avg_conditional_strength = replace_na(avg_conditional_strength, 0),
    n_pairs = replace_na(n_pairs, 0)
  )

# baseline score
case8_basket <- case8_basket %>%
  mutate(
    fast_scan = avg_time_per_item < p5_time,
    slow_scan = avg_time_per_item > p95_time,
    low_total_price = total_price < p5_price,
    high_total_price = total_price > p95_price,
    unusually_cheap_for_size = price_deviation < p5_resid,
    unusually_expensive_for_size = price_deviation > p95_resid,
    high_price_concentration = price_concentration > p95_conc,
    unusual_department_combination = co_occurrence_score < p5_co,
    risk_score =
      as.integer(fast_scan) +
      as.integer(slow_scan) +
      as.integer(low_total_price) +
      as.integer(high_total_price) +
      as.integer(unusually_cheap_for_size) +
      as.integer(unusually_expensive_for_size) +
      as.integer(high_price_concentration) +
      as.integer(unusual_department_combination),
    final_score =
      risk_score * 2 +
      as.numeric(scale(total_price)) +
      as.numeric(scale(abs(price_deviation))) +
      as.numeric(scale(abs(time_deviation))) * 0.5
  )

# IF score
if_case8 <- case8_basket %>%
  select(
    transaction_id,
    total_items,
    total_price,
    avg_price,
    total_time,
    avg_time_per_item,
    sd_time_per_item,
    price_per_second,
    items_per_second,
    max_price,
    max_price_share,
    unique_departments,
    department_diversity_ratio,
    dominant_department_share,
    price_deviation,
    time_deviation
  ) %>%
  mutate(
    sd_time_per_item = replace_na(sd_time_per_item, median(if_train$sd_time_per_item, na.rm = TRUE)),
    price_per_second = replace_na(price_per_second, median(if_train$price_per_second, na.rm = TRUE)),
    items_per_second = replace_na(items_per_second, median(if_train$items_per_second, na.rm = TRUE)),
    max_price_share = replace_na(max_price_share, median(if_train$max_price_share, na.rm = TRUE)),
    department_diversity_ratio = replace_na(department_diversity_ratio, median(if_train$department_diversity_ratio, na.rm = TRUE)),
    dominant_department_share = replace_na(dominant_department_share, median(if_train$dominant_department_share, na.rm = TRUE)),
    price_deviation = replace_na(price_deviation, median(if_train$price_deviation, na.rm = TRUE)),
    time_deviation = replace_na(time_deviation, median(if_train$time_deviation, na.rm = TRUE))
  )

X_case8_raw <- if_case8 %>%
  select(-transaction_id)

X_case8 <- scale(
  X_case8_raw,
  center = attr(X_train, "scaled:center"),
  scale = attr(X_train, "scaled:scale")
)

if_case8$if_score <- predict(
  if_model,
  as.matrix(X_case8),
  type = "score"
)

# final combined score
case8_big <- case8_basket %>%
  left_join(
    if_case8 %>% select(transaction_id, if_score),
    by = "transaction_id"
  ) %>%
  mutate(
    baseline_score_01 = percent_rank(final_score),
    if_score_01 = percent_rank(if_score),
    graph_score_01 = percent_rank(graph_score),
    combined_big_score =
      0.25 * baseline_score_01 +
      0.45 * if_score_01 +
      0.30 * graph_score_01
  ) %>%
  arrange(desc(combined_big_score))

top_150_case8_big <- case8_big %>%
  slice_head(n = 150)

summary(top_150_case8_big$final_score)
summary(top_150_case8_big$if_score)
summary(top_150_case8_big$graph_score)

writeLines(as.character(top_150_case8_big$transaction_id))

readr::write_csv(
  top_150_case8_big,
  "~/Desktop/supermarket case/test cases/top_150_case8_baseline_if_graph.csv"
)
```



# OPTUNA!!!!! КАЛИБРОВКА ВЕСОВ


```{r}
library(reticulate)

use_virtualenv("~/.virtualenvs/r-reticulate", required = TRUE)
optuna <- import("optuna")

names(baseline_train)
```

```{r}
objective <- function(trial) {
  w_baseline <- trial$suggest_float("w_baseline", 0.0, 0.4)
  w_if       <- trial$suggest_float("w_if",       0.2, 0.7)
  w_graph    <- trial$suggest_float("w_graph",    0.1, 0.5)

  top_k <- trial$suggest_categorical("top_k", list(100L, 125L, 150L, 175L, 200L))

  w_sum <- w_baseline + w_if + w_graph
  w_baseline <- w_baseline / w_sum
  w_if <- w_if / w_sum
  w_graph <- w_graph / w_sum

  scored <- baseline_train %>%
    mutate(
      optuna_score =
        w_baseline * baseline_score_01 +
        w_if * if_score_01 +
        w_graph * graph_score_01
    ) %>%
    arrange(desc(optuna_score))

  top_data <- scored %>%
    slice_head(n = top_k)

  mean_items <- mean(top_data$total_items, na.rm = TRUE)
  mean_price <- mean(top_data$total_price, na.rm = TRUE)

  # простая нормализация штрафов
  items_penalty <- mean_items / max(baseline_train$total_items, na.rm = TRUE)
  price_penalty <- mean_price / max(baseline_train$total_price, na.rm = TRUE)

  rule_bonus <- mean(
    as.numeric(top_data$high_price_concentration) +
    as.numeric(top_data$unusual_department_combination) +
    as.numeric(top_data$slow_scan),
    na.rm = TRUE
  ) / 3

  objective_value <-
    0.35 * mean(top_data$if_score_01, na.rm = TRUE) +
    0.30 * mean(top_data$graph_score_01, na.rm = TRUE) +
    0.20 * mean(top_data$baseline_score_01, na.rm = TRUE) +
    0.15 * rule_bonus -
    0.03 * items_penalty -
    0.03 * price_penalty

  return(objective_value)
}

study <- optuna$create_study(direction = "maximize")
study$optimize(objective, n_trials = as.integer(100))
```


```{r}
study$best_params
study$best_value
```


```{r}
best_params <- study$best_params

w_baseline <- best_params[["w_baseline"]]
w_if <- best_params[["w_if"]]
w_graph <- best_params[["w_graph"]]

w_sum <- w_baseline + w_if + w_graph
w_baseline <- w_baseline / w_sum
w_if <- w_if / w_sum
w_graph <- w_graph / w_sum

top_k_best <- as.integer(best_params[["top_k"]])

final_optuna_train <- baseline_train %>%
  mutate(
    optuna_score =
      w_baseline * baseline_score_01 +
      w_if * if_score_01 +
      w_graph * graph_score_01
  ) %>%
  arrange(desc(optuna_score))

top_optuna_train <- final_optuna_train %>%
  slice_head(n = top_k_best)

summary(top_optuna_train$if_score)
summary(top_optuna_train$graph_score)
summary(top_optuna_train$total_items)
summary(top_optuna_train$total_price)
```


```{r}
write.csv(
  top_optuna_train,
  "~/Desktop/supermarket case/test cases/top_optuna_train.csv",
  row.names = FALSE
)
```

# OPTUNA WEIGHTS SECOND TEST

```{r}
objective <- function(trial) {
  w_baseline <- trial$suggest_float("w_baseline", 0.05, 0.45)
  w_if       <- trial$suggest_float("w_if", 0.20, 0.70)
  w_graph    <- trial$suggest_float("w_graph", 0.10, 0.50)

  top_k <- as.integer(
    trial$suggest_categorical("top_k", list(100L, 125L, 150L, 175L, 200L))
  )

  w_sum <- w_baseline + w_if + w_graph
  w_baseline <- w_baseline / w_sum
  w_if <- w_if / w_sum
  w_graph <- w_graph / w_sum

  scored <- baseline_train %>%
    mutate(
      optuna_score =
        w_baseline * baseline_score_01 +
        w_if * if_score_01 +
        w_graph * graph_score_01
    ) %>%
    arrange(desc(optuna_score))

  top_data <- scored %>%
    slice_head(n = top_k)

  mean_items <- mean(top_data$total_items, na.rm = TRUE)
  mean_price <- mean(top_data$total_price, na.rm = TRUE)

  median_items <- median(top_data$total_items, na.rm = TRUE)
  median_price <- median(top_data$total_price, na.rm = TRUE)

  items_penalty <- mean_items / max(baseline_train$total_items, na.rm = TRUE)
  price_penalty <- mean_price / max(baseline_train$total_price, na.rm = TRUE)

  rule_bonus <- mean(
    as.numeric(top_data$high_price_concentration) +
    as.numeric(top_data$unusual_department_combination) +
    as.numeric(top_data$slow_scan),
    na.rm = TRUE
  ) / 3

  collapse_penalty <- 0

  if (median_items < 5) {
    collapse_penalty <- collapse_penalty + 0.05
  }

  if (median_price < 15) {
    collapse_penalty <- collapse_penalty + 0.05
  }

  objective_value <-
    0.33 * mean(top_data$if_score_01, na.rm = TRUE) +
    0.28 * mean(top_data$graph_score_01, na.rm = TRUE) +
    0.24 * mean(top_data$baseline_score_01, na.rm = TRUE) +
    0.15 * rule_bonus -
    0.03 * items_penalty -
    0.03 * price_penalty -
    collapse_penalty

  return(objective_value)
}
```

```{r}
study <- optuna$create_study(direction = "maximize")
study$optimize(objective, n_trials = as.integer(100))
```


```{r}
study$best_params
study$best_value
```


```{r}
best_params <- study$best_params

w_baseline <- best_params[["w_baseline"]]
w_if <- best_params[["w_if"]]
w_graph <- best_params[["w_graph"]]

w_sum <- w_baseline + w_if + w_graph
w_baseline <- w_baseline / w_sum
w_if <- w_if / w_sum
w_graph <- w_graph / w_sum

top_k_best <- as.integer(best_params[["top_k"]])

final_optuna_train <- baseline_train %>%
  mutate(
    optuna_score =
      w_baseline * baseline_score_01 +
      w_if * if_score_01 +
      w_graph * graph_score_01
  ) %>%
  arrange(desc(optuna_score))

top_optuna_train <- final_optuna_train %>%
  slice_head(n = top_k_best)

summary(top_optuna_train$if_score)
summary(top_optuna_train$graph_score)
summary(top_optuna_train$total_items)
summary(top_optuna_train$total_price)

write.csv(
  top_optuna_train,
  "~/Desktop/supermarket case/test cases/top_optuna_train_v2.csv",
  row.names = FALSE
)
```
# OPTUNA 3RD TEST

```{r}
objective <- function(trial) {
  w_baseline <- trial$suggest_float("w_baseline", 0.05, 0.50)
  w_if       <- trial$suggest_float("w_if", 0.15, 0.70)
  w_graph    <- trial$suggest_float("w_graph", 0.10, 0.50)

top_k <- 150L

  w_sum <- w_baseline + w_if + w_graph
  w_baseline <- w_baseline / w_sum
  w_if <- w_if / w_sum
  w_graph <- w_graph / w_sum

  scored <- baseline_train %>%
    mutate(
      optuna_score =
        w_baseline * baseline_score_01 +
        w_if * if_score_01 +
        w_graph * graph_score_01
    ) %>%
    arrange(desc(optuna_score))

  top_data <- scored %>%
    slice_head(n = top_k)

  mean_items <- mean(top_data$total_items, na.rm = TRUE)
  mean_price <- mean(top_data$total_price, na.rm = TRUE)

  median_items <- median(top_data$total_items, na.rm = TRUE)
  median_price <- median(top_data$total_price, na.rm = TRUE)

  # hard constraints
  if (median_items < 5) {
    return(-999)
  }

  if (median_price < 15) {
    return(-999)
  }

  if (mean_items < 6) {
    return(-999)
  }

  if (mean_price < 18) {
    return(-999)
  }

  items_penalty <- mean_items / max(baseline_train$total_items, na.rm = TRUE)
  price_penalty <- mean_price / max(baseline_train$total_price, na.rm = TRUE)

  rule_bonus <- mean(
    as.numeric(top_data$high_price_concentration) +
    as.numeric(top_data$unusual_department_combination) +
    as.numeric(top_data$slow_scan),
    na.rm = TRUE
  ) / 3

  objective_value <-
    0.32 * mean(top_data$if_score_01, na.rm = TRUE) +
    0.26 * mean(top_data$graph_score_01, na.rm = TRUE) +
    0.27 * mean(top_data$baseline_score_01, na.rm = TRUE) +
    0.15 * rule_bonus -
    0.02 * items_penalty -
    0.02 * price_penalty

  return(objective_value)
}
```


```{r}
study <- optuna$create_study(direction = "maximize")
study$optimize(objective, n_trials = as.integer(150))
```


```{r}
study$best_params
study$best_value
```

```{r}
 best_params <- study$best_params

w_baseline <- best_params[["w_baseline"]]
w_if <- best_params[["w_if"]]
w_graph <- best_params[["w_graph"]]

w_sum <- w_baseline + w_if + w_graph
w_baseline <- w_baseline / w_sum
w_if <- w_if / w_sum
w_graph <- w_graph / w_sum

top_k_best <- 150L

final_optuna_train <- baseline_train %>%
  mutate(
    optuna_score =
      w_baseline * baseline_score_01 +
      w_if * if_score_01 +
      w_graph * graph_score_01
  ) %>%
  arrange(desc(optuna_score))

top_optuna_train <- final_optuna_train %>%
  slice_head(n = top_k_best)

summary(top_optuna_train$if_score)
summary(top_optuna_train$graph_score)
summary(top_optuna_train$total_items)
summary(top_optuna_train$total_price)

write.csv(
  top_optuna_train,
  "~/Desktop/supermarket case/test cases/top_optuna_train_v3.csv",
  row.names = FALSE
)
```


```{r}

```
