---
title: "Pattern EDA"
output:
  html_document: default
  pdf_document: default
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
long_data <- read.csv("C:/Users/nicol/OneDrive/universita/elab 2/long_data.csv", stringsAsFactors = FALSE)
```

```{r}
#REMOVE DUPLICATES WITHIN TRANSACTION
dept_data <- long_data %>%
  select(transaction_id, department_name) %>%
  distinct()

#CREATE CO-OCCURRENCE PAIRS
pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y)

#COUNT PAIRS
co_occurrence <- pairs %>%
  count(department_name.x, department_name.y, sort = TRUE)

#TOP CO-OCCURRENCE PATTERNS
head(co_occurrence, 10)
```
After removing duplicate departments within each transaction, we generated all possible pairs of co-occurring categories and counted their frequency across all baskets. The output shows that combinations involving Fruit & Vegetables dominate, reflecting its high presence in transactions, while other frequent pairs such as Pasta & Rice with Sauces & Spices highlight more structured and complementary purchasing behavior.

```{r}
#CO-OCCURRENCE PER DEPARTMENT
co_by_dept <- co_occurrence %>%
  group_by(department_name.x) %>%
  slice_max(n, n = 4) %>%
  ungroup()

co_by_dept
```
The output shows, for each department, the most frequent co-occurring categories within the same transaction. While some combinations are driven by highly frequent categories such as Fruit & Vegetables, other patterns appear more structured and meaningful, particularly those involving Pasta & Rice, Sauces & Spices, and Meat & Fish, which frequently co-occur across multiple departments. These results suggest that customers tend to purchase complementary products rather than independent items, reflecting consistent and non-random shopping behavior.

```{r}
#TOTAL NUMBER OF TRANSACTIONS
total_transactions <- length(unique(long_data$transaction_id))

#FREQUENCY OF EACH DEPARTMENT (PERCENTAGE OF CUSTOMERS)
dept_freq <- dept_data %>%
  count(department_name) %>%
  mutate(
    perc_total = n / total_transactions
  )

#JOIN CO-OCCURRENCE WITH DEPARTMENT FREQUENCIES
co_occurrence_perc <- co_occurrence %>%
  left_join(dept_freq, by = c("department_name.x" = "department_name"))

#RENAME COLUMNS AND COMPUTE CONDITIONAL PROBABILITY
co_occurrence_perc <- co_occurrence_perc %>%
  rename(
    pair_count = n.x,
    dept_count = n.y,
    dept_share = perc_total
  ) %>%
  mutate(
    conditional_prob = pair_count / dept_count
  )

#VIEW TOP STRONGEST RELATIONSHIPS
co_occurrence_perc %>%
  arrange(desc(conditional_prob)) %>%
  head(10)
```
The results show the strongest relationships between departments in terms of conditional probability, highlighting how likely a category is to be purchased given another. While combinations involving Fruit & Vegetables dominate due to its high overall presence, other relationships such as Dairy & Eggs with Sauces & Spices or Cereals & Spreads with Sauces & Spices reveal more meaningful and structured purchasing behavior, indicating that customers tend to select complementary products rather than random combinations.

```{r}
#TOP CO-OCCURRENCE PER EACH DEPARTMENT
co_by_dept <- co_occurrence_perc %>%
  group_by(department_name.x) %>%
  slice_max(conditional_prob, n = 5) %>%
  ungroup()

co_by_dept
```

```{r}
#DEFINE EXCLUDED DEPARTMENTS
excluded_depts <- c(
  "Fruit & Vegetables",
  "Books & Magazines",
  "Care & Hygiene",
  "Household & Pet"
)

#FILTER AND GET TOP 10 RELATIONSHIPS
top10_filtered <- co_occurrence_perc %>%
  filter(
    !department_name.x %in% excluded_depts,
    !department_name.y %in% excluded_depts
  ) %>%
  arrange(desc(conditional_prob)) %>%
  slice_head(n = 10)

top10_filtered
```
After excluding dominant and non-core categories, the results highlight clearer and more meaningful relationships between departments. Strong associations emerge between staple and complementary products, such as Pasta & Rice with Sauces & Spices and Dairy & Eggs with Sauces & Spices, indicating structured purchasing behavior driven by meal preparation. These patterns are more informative than the unfiltered results and provide a more reliable baseline for identifying unusual or potentially anomalous transactions.


From the analysis, especially from the analysis based on probabilities and from the filtered analysis, some patterns emerge that appear to be random, such as Books & Magazines or Care & Hygiene with Fruit & Vegetables. These are likely due to the high number of transactions that include Fruit & Vegetables. At the same time, other combinations appear more meaningful and could be more relevant to investigate when searching for potentially fraudulent customers.
1. Dairy & Eggs → Sauces & Spices (~80%)
2. Cereals & Spreads → Sauces (~77%)
3. Pasta & Rice → Sauces (~71%)
4. Cheese & Tapas → Pasta (~74%)
5. Meat & Fish → Pasta (~68%)