---
title: "Pattern analysis"
output: html_document
date: "2026-04-19"
---
```{r}
long_data <- read.csv("C:/Users/nicol/OneDrive/universita/elab 2/long_data.csv", stringsAsFactors = FALSE)
```

```{r}
#REMOVE DUPLICATES WITHIN TRANSACTION
dept_data <- long_data %>%
  select(transaction_id, department_name) %>%
  distinct()

#CREATE CO-OCCURRENCE PAIRS
pairs <- dept_data %>%
  inner_join(dept_data, by = "transaction_id") %>%
  filter(department_name.x < department_name.y)

#COUNT PAIRS
co_occurrence <- pairs %>%
  count(department_name.x, department_name.y, sort = TRUE)

#TOP CO-OCCURRENCE PATTERNS
head(co_occurrence, 10)
```

```{r}
#CO-OCCURRENCE PER DEPARTMENT
co_by_dept <- co_occurrence %>%
  group_by(department_name.x) %>%
  slice_max(n, n = 10) %>%
  ungroup()

co_by_dept
```

```{r}
#TOTAL NUMBER OF TRANSACTIONS
total_transactions <- length(unique(long_data$transaction_id))

#FREQUENCY OF EACH DEPARTMENT (PERCENTAGE OF CUSTOMERS)
dept_freq <- dept_data %>%
  count(department_name) %>%
  mutate(
    perc_total = n / total_transactions
  )

#JOIN CO-OCCURRENCE WITH DEPARTMENT FREQUENCIES
co_occurrence_perc <- co_occurrence %>%
  left_join(dept_freq, by = c("department_name.x" = "department_name"))

#RENAME COLUMNS AND COMPUTE CONDITIONAL PROBABILITY
co_occurrence_perc <- co_occurrence_perc %>%
  rename(
    pair_count = n.x,
    dept_count = n.y,
    dept_share = perc_total
  ) %>%
  mutate(
    conditional_prob = pair_count / dept_count
  )

#VIEW TOP STRONGEST RELATIONSHIPS
co_occurrence_perc %>%
  arrange(desc(conditional_prob)) %>%
  head(10)
```

```{r}
#TOP CO-OCCURRENCE PER EACH DEPARTMENT
co_by_dept <- co_occurrence_perc %>%
  group_by(department_name.x) %>%
  slice_max(conditional_prob, n = 10) %>%
  ungroup()

co_by_dept
```

```{r}
#DEFINE EXCLUDED DEPARTMENTS
excluded_depts <- c(
  "Fruit & Vegetables",
  "Books & Magazines",
  "Care & Hygiene",
  "Household & Pet"
)

#FILTER AND GET TOP 10 RELATIONSHIPS
top10_filtered <- co_occurrence_perc %>%
  filter(
    !department_name.x %in% excluded_depts,
    !department_name.y %in% excluded_depts
  ) %>%
  arrange(desc(conditional_prob)) %>%
  slice_head(n = 10)

top10_filtered
```

From the analysis, especially from the analysis based on probabilities and from the filtered analysis, some patterns emerge that appear to be random, such as Books & Magazines or Care & Hygiene with Fruit & Vegetables. These are likely due to the high number of transactions that include Fruit & Vegetables. At the same time, other combinations appear more meaningful and could be more relevant to investigate when searching for potentially fraudulent customers.
1. Dairy & Eggs → Sauces & Spices (~80%)
2. Cereals & Spreads → Sauces (~77%)
3. Pasta & Rice → Sauces (~71%)
4. Cheese & Tapas → Pasta (~74%)
5. Meat & Fish → Pasta (~68%)

